---
title: "Sparse NCA Calculations"
author: "Bill Denney"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Sparse NCA Calculations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Sparse NCA Calculations

Sparse noncompartmental analysis (NCA) is performed when multiple individuals contribute to a single concentration-time profile due to the fact that there are only one or a subset of the full profile samples taken per animal.  A typical example is when three mice have PK drawn per time point, but no animals have more than one sample drawn.  Another typical example is when animals may have two or three samples during an interval, but no animal has the full profile.

## Sparse NCA Setup

Sparse NCA is setup the same way as normal, dense PK sampling is setup with PKNCA.  The only difference is that you give the `sparse` option to `PKNCAconc()`; the parameters are requested by their usual names.

Two things happen behind that flag.  For the AUC and AUMC parameters with a sparse estimator, PKNCA calculates them from the pooled samples of all the animals in a group with the Bailer point estimate and the Nedelman-Jia/Holder standard error, and reports the standard error and degrees of freedom alongside the estimate with the suffixes `_se` and `_df` (for example, `auclast_se` and `auclast_df`).  Those parameters are `auclast`, `aucall`, `aucinf.obs`, `aumclast`, `aumcall`, and `aumcinf.obs`, and after an IV bolus `aucivlast`, `aucivall`, `aucivinf.obs`, `aumcivlast`, `aumcivall`, and `aumcivinf.obs` (see `vignette("v24-sparse-auc-to-infinity")` for the AUC to infinity and the IV bolus $C_0$).  Every other parameter is calculated from the arithmetic-mean profile of the animals in the group.

As for any AUC in PKNCA, the sparse AUCs need a concentration at the start of the interval, measured or imputed.  Animals are rarely sampled at the dose in sacrificial designs, so after an extravascular dose impute zero with `impute = "start_conc0"`; the imputed zero is known, so it adds nothing to the variance.

The example below uses data extracted from Holder D. J., Hsuan F., Dixit R. and Soper K. (1999). A method for estimating and testing area under the curve in serial sacrifice, batch, and complete data designs. Journal of Biopharmaceutical Statistics, 9(3):451-464.

```{r}
# Setup the data
d_sparse <-
    data.frame(
      id = c(1L, 2L, 3L, 1L, 2L, 3L, 1L, 2L, 3L, 4L, 5L, 6L, 4L, 5L, 6L, 7L, 8L, 9L, 7L, 8L, 9L),
      conc = c(0, 0, 0,  1.75, 2.2, 1.58, 4.63, 2.99, 1.52, 3.03, 1.98, 2.22, 3.34, 1.3, 1.22, 3.54, 2.84, 2.55, 0.3, 0.0421, 0.231),
      time = c(0, 0, 0, 1, 1, 1, 6, 6, 6, 2, 2, 2, 10, 10, 10, 4, 4, 4, 24, 24, 24),
      dose = c(100, 100, 100, 100, 100, 100, 100, 100, 100, 100, 100, 100, 100, 100, 100, 100, 100, 100, 100, 100, 100)
    )
```

Look at your data.  (This is not technically a required step, but it's good practice.)

```{r}
library(ggplot2)
ggplot(d_sparse, aes(x=time, y=conc, group=id)) +
  geom_point() +
  geom_line() +
  scale_x_continuous(breaks=seq(0, 24, by=6))
```

## Data Setup Note

Sparse NCA requires that subject numbers (or animal numbers) are given, even if each subject only contributes a single sample.  The reason for this requirement is that which subject contributes to which time point changes the standard error calculation.  If all individuals contribute a single sample, a simple way to handle this is by setting a column with sequential numbers and giving that as the subject identifier:

```{r eval=FALSE}
d_sparse$id <- 1:nrow(d_sparse)
```

## How Subjects Are Grouped for Sparse Calculations

With dense (normal) PK, every subject has a full concentration-time profile, so NCA parameters are calculated one subject at a time.  Sparse parameters are different: they are calculated from the *pooled* samples of every subject that belongs to the same group.  Knowing what defines a "group" is therefore important.

The groups are taken from the grouping variables on the right of the `|` in the `PKNCAconc()` formula, **with the subject column removed**.  Every subject that shares the same combination of the remaining (non-subject) grouping variables contributes to a single pooled sparse concentration-time profile.

In the simple example above, the formula is `conc~time|id`.  Here `id` is the subject, and removing it leaves no other grouping variables, so all of the data form a single sparse group.

The behavior is easier to see with more grouping variables.  Suppose the concentration and dose objects are created with the formulas below (illustrative code; not run here):

```{r eval=FALSE}
o_conc_sparse <- PKNCAconc(d_conc, conc~time|matrix+drug+usubjid/analyte, sparse=TRUE)
o_dose_sparse <- PKNCAdose(d_dose, dose~time|drug+usubjid)
```

`usubjid` is the subject because, by default, the subject is the last grouping variable before any `/` (or the last grouping variable when there is no `/`).  After dropping the subject, the grouping variables that remain are `matrix`, `drug`, and `analyte`.  Sparse parameters are therefore calculated by combining all subjects within each unique combination of `matrix`, `drug`, and `analyte`.

In other words, with that formula PKNCA does **not** keep each `usubjid` separate, and it does **not** group by `matrix`, `drug`, or `analyte` alone.  It pools subjects using the full set of non-subject grouping variables together (`matrix` + `drug` + `analyte`).

Because subjects are pooled within a group, all subjects in a group must share the same dosing.  If subjects in the same group have different dosing information (for example, different dose amounts or dose times), PKNCA stops with an error identifying the inconsistent group.

# Calculate!

Setup PKNCA for calculations and then calculate!

```{r}
library(PKNCA)
o_conc_sparse <- PKNCAconc(d_sparse, conc~time|id, sparse=TRUE)
d_intervals <-
  data.frame(
    start=0,
    end=24,
    auclast=TRUE,
    auclast_se=TRUE,
    auclast_df=TRUE,
    aucinf.obs=TRUE,
    cmax=TRUE
  )
o_data_sparse <- PKNCAdata(o_conc_sparse, intervals=d_intervals)
o_nca <- pk.nca(o_data_sparse)
```

`pknca_interval_table()` builds the same kind of specification from a description of the study, and its `sparse_single_dose` preset is the single-dose set with a zero imputed at the start (`start_conc0`), the one imputation the sparse estimators accept:

```{r eval=FALSE}
pknca_interval_table(0, 24, preset="sparse_single_dose")
```

# Results

As with any other PKNCA result, the data are available through the `summary()` function:

```{r}
summary(o_nca)
```

In the summary, the sparse `auclast` is the estimate with its standard error in brackets (from the `auclast_se` result), as the caption says; `auclast_se` has no column of its own.

or individual results are available through the `as.data.frame()` function:

```{r}
as.data.frame(o_nca)
```

`auclast_se` and `auclast_df` are reported whether or not they were requested, because the sparse estimator returns all three together.  They are only calculable from sparse data:  requesting one for a dense `PKNCAconc` is an error.

# Sparse AUC, AUMC, and Derived Parameters

Because the sparse estimate is stored under the standard parameter name, every parameter derived from an AUC is available for sparse data with no extra work.  `cl.last` divides the dose by `auclast`, `mrt.last` divides `aumclast` by `auclast`, and so on down the usual derived graph.

```{r sparse-derived}
d_dose <- data.frame(id=unique(d_sparse$id), dose=100, time=0)
o_dose <- PKNCAdose(d_dose, dose~time|id, route="intravascular")
d_intervals_derived <-
  data.frame(
    start=0,
    end=24,
    auclast=TRUE,
    aumclast=TRUE,
    mrt.last=TRUE,
    cl.last=TRUE,
    kel.last=TRUE,
    vss.last=TRUE,
    vz.last=TRUE
  )
o_data_derived <- PKNCAdata(o_conc_sparse, o_dose, intervals=d_intervals_derived)
o_nca_derived <- pk.nca(o_data_derived)
as.data.frame(o_nca_derived)
```

Note what `vz.last` does here.  It is the clearance divided by $\lambda_z$, the terminal rate constant fitted on the mean profile, which is the standard toxicokinetic approach.  These data have only two time points after the mean profile's peak, which is too few for a terminal fit, so $\lambda_z$ and therefore `vz.last` are `NA`.  That is the honest answer:  the retired `vz.sparse.last` divided the clearance by `1/MRT` instead, which always produced a number but made Vz numerically identical to Vss.

`kel.last`, like the rest of PKNCA's `kel` family, is `1/MRT`, so it is unchanged from the retired `kel.sparse.last`.  Ask for `lambda.z` when you want the terminal rate constant itself.

# Bootstrap

The sparse standard errors above come from the variance of the means at each time.  A nonparametric bootstrap (Shen and Machado 2017) is an alternative that applies to every parameter, including Cmax, the half-life, and ratios between groups:  `sparse_bootstrap()` resamples the animals with replacement within each set of sampling times (each time with serial sacrifice, each batch in a batch design, keeping an animal's samples together), and every replicate is calculated just as the original data are.  It stores the random seed so the replicates can be reproduced, and the default of 200 replicates is enough for the mean and standard deviation (Takemoto et al. 2006).

```{r bootstrap}
o_conc_boot <- sparse_bootstrap(o_conc_sparse, seed=20261007)
o_data_boot <-
  PKNCAdata(
    o_conc_boot,
    intervals=data.frame(start=0, end=24, auclast=TRUE, cmax=TRUE, tmax=TRUE),
    # Many replicate groups would otherwise show a progress bar
    options=list(progress=FALSE)
  )
o_nca_boot <- pk.nca(o_data_boot)
sparse_bootstrap_summary(o_nca_boot)
```

For each parameter, `estimate` is the value from the original data, and `mean`, `sd`, `geomean`, and `geocv` describe the replicates; the bootstrap standard deviation is the standard error of the estimate, and `ci_lower` and `ci_upper` are the percentile interval.  With only three animals at each time, the bootstrap standard deviation of `auclast` is smaller than its Nedelman-Jia/Holder standard error (`auclast_se` above), because resampling three values understates their variance by a factor of 2/3; `vignette("v24-sparse-auc-to-infinity")` compares the methods.

To compare groups, such as a test and a reference formulation in a parallel design, `sparse_bootstrap_compare()` gives the ratio of the estimates with the percentile interval of the replicate ratios (90% by default) and whether it is within the acceptance limits.  For a crossover, where each animal receives every treatment, give the treatment column as `paired` to `sparse_bootstrap()` so that each replicate draws the same animals for every treatment.

```{r bootstrap-compare}
d_sparse_formulations <-
  rbind(
    data.frame(d_sparse, formulation="reference"),
    data.frame(
      id=d_sparse$id + 100, conc=d_sparse$conc*c(0.9, 1.05, 1.1), time=d_sparse$time,
      dose=d_sparse$dose, formulation="test"
    )
  )
o_conc_formulations <- PKNCAconc(d_sparse_formulations, conc~time|formulation+id, sparse=TRUE)
o_nca_formulations <-
  pk.nca(PKNCAdata(
    sparse_bootstrap(o_conc_formulations, seed=20261008),
    intervals=data.frame(start=0, end=24, auclast=TRUE, cmax=TRUE),
    options=list(progress=FALSE)
  ))
sparse_bootstrap_compare(
  o_nca_formulations, reference_col="formulation", reference_value="reference"
)
```

The bootstrap methods are from Shen M. and Machado S. G. (2017). Bioequivalence evaluation of sparse sampling pharmacokinetics data using bootstrap resampling method. Journal of Biopharmaceutical Statistics, 27(2):257-264, and Takemoto S., Yamaoka K., Nishikawa M. and Takakura Y. (2006). Histogram analysis of pharmacokinetic parameters by bootstrap resampling from one-point sampling data in animal experiments. Drug Metabolism and Pharmacokinetics, 21(6):458-464.

# Deprecated Parameter Names

Sparse calculations were originally requested through a parallel set of names.  All eleven still calculate and still give the values they always gave, but each now warns and **will be an error in the next minor release of PKNCA**:

| Deprecated | Use instead |
|---|---|
| `sparse_auclast` | `auclast` |
| `sparse_auc_se` | `auclast_se` |
| `sparse_auc_df` | `auclast_df` |
| `sparse_aumclast` | `aumclast` |
| `sparse_aumc_se` | `aumclast_se` |
| `sparse_aumc_df` | `aumclast_df` |
| `cl.sparse.last` | `cl.last` |
| `mrt.sparse.last` | `mrt.last` |
| `kel.sparse.last` | `kel.last` |
| `vss.sparse.last` | `vss.last` |
| `vz.sparse.last` | `vz.last` |

Every replacement but the last gives the same number as the name it replaces.  `vz.last` is $\lambda_z$-based, as described above, so it differs from `vz.sparse.last` by design.

The functions behind the sparse estimators -- `pk.calc.sparse_auc()`, `pk.calc.sparse_auclast()`, `pk.calc.sparse_aumc()`, `pk.calc.sparse_aumclast()`, `as_sparse_pk()`, `sparse_mean()`, `var_sparse_auc()`, and `cov_holder()` -- are not deprecated.  They are what the unified estimators call, and they remain available for calculations outside `pk.nca()`.

# Notes on Sparse Calculation Behavior

## Degrees of freedom with multiple samples per subject

The degrees of freedom (`auclast_df` and `aumclast_df`) are the Satterthwaite approximation of Nedelman and Jia (1998), which accounts for the correlation between samples from the same subject, so they are calculated whether each subject contributes one sample (as in a serial sacrifice design) or several (as in the example data here and in batch designs).  They are `NA` when a time point has a single subject, because its variance cannot be estimated.

## More than half of the measurements below the limit of quantification

When calculating the mean concentration at a time point (see `sparse_mean()`), if strictly more than 50% of the measurements at that time point are below the limit of quantification (BLQ), the mean for that time point is set to zero.  At exactly 50% BLQ, the mean is calculated normally (including the BLQ values as zero).

## Only the linear trapezoidal method is supported

The sparse variance theory is derived for a fixed-weight linear combination of the per-time-point means, so the sparse estimators are defined with the linear trapezoidal method only.  Calling `pk.calc.sparse_auc()` with any other `method` is an error, and within `pk.nca()` the sparse estimators always use the linear method:  the `auc.method` option does not apply to them, and `pk.nca()` says so once per run when the option asks for anything else.  Parameters that fall back to the mean profile, such as `aucinf.obs`, do follow `auc.method`.
