---
title: "AUC Calculation with PKNCA"
author: "Bill Denney"
output:
  rmarkdown::html_vignette:
    toc: yes
    toc_depth: 6
vignette: >
  %\VignetteIndexEntry{AUC Calculation with PKNCA}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

Area Under the Curve (AUC) is a commonly-used metric for assessing exposure to a drug.  Many variants of AUC exist, and the information below will assist in determining both the appropriate AUC and how to calculate it.

```{r check-ggplot, include=!requireNamespace("ggplot2"), results="asis"}
cat("ggplot2 is required for this vignette to work correctly.  Please install the ggplot2 library and retry building the vignette.")
```

# Preparation

For the below examples, the following data will be used.

```{r setup}
suppressPackageStartupMessages({
  library(PKNCA)
  library(dplyr)
  library(cowplot)
  library(knitr)
  library(ggplot2)
})
scale_colour_discrete <- scale_colour_hue
scale_fill_discrete <- scale_fill_hue

my_conc <- data.frame(conc=c(0, 2.5, 3, 2, 1.5, 1.2, 1.1, 0, 0),
                      time=c(0:5, 8, 12, 24),
                      subject=1)
my_conc$BLQ <- my_conc$conc == 0
my_conc$measured <- TRUE
```
```{r setup-visualization, eval=requireNamespace("ggplot2")}
ggplot(my_conc,
       aes(x=time,
           y=conc,
           shape=BLQ,
           group=subject)) +
  geom_line() +
  geom_point(size=4) +
  scale_x_continuous(breaks=my_conc$time) +
  theme(legend.position=c(0.8, 0.8))
```

For the purpose of illustration, NCA parameters will also be calculated for each of the AUC types below.  Note that in the results, more parameters are returned than were requested.  The additional parameters are the set of parameters required to calculate the requested parameters.

```{r calc_nca}
conc_obj <- PKNCAconc(my_conc, conc~time|subject)
data_obj <- PKNCAdata(data.conc=conc_obj,
                      intervals=data.frame(start=0,
                                           end=24,
                                           aucall=TRUE,
                                           auclast=TRUE,
                                           aucinf.pred=TRUE,
                                           aucinf.obs=TRUE))
results_obj <- pk.nca(data_obj)
kable(as.data.frame(results_obj))
```

# AUC to the Last Value Above the Limit of Quantification (AUC~last~)

AUC~0-last~ calculates the AUC from time 0 to the last value above the limit of quantification, `tlast` (within PKNCA, this is the last value above 0).  In the figure below, AUC~0-last~ integrates the shaded region.  Integration after `tlast` is 0.

```{r auclast}
tlast <- pk.calc.tlast(conc=my_conc$conc,
                       time=my_conc$time)
tlast

my_conc$include_auclast <- my_conc$time <= tlast
```
```{r auclast-visualization, eval=requireNamespace("ggplot2")}
ggplot(my_conc,
       aes(x=time,
           y=conc,
           shape=BLQ,
           group=subject)) +
  geom_ribbon(data=my_conc[my_conc$include_auclast,],
              aes(ymin=0, ymax=conc),
              fill="lightblue") +
  geom_line() +
  geom_point(size=4) +
  scale_x_continuous(breaks=my_conc$time) +
  theme(legend.position=c(0.8, 0.8))
```

# AUC~all~

AUC~all~ starts with AUC~0-last~ and then integrates from `tlast` to the first point after `tlast` with a linear interpolation to zero.  From the second point after `tlast` to $\infty$ is considered zero.

```{r aucall}
first_after_tlast <- my_conc$time[my_conc$time > tlast][1]
first_after_tlast

my_conc$include_aucall <- my_conc$time <= first_after_tlast
```
```{r aucall-visualization, eval=requireNamespace("ggplot2")}
ggplot(my_conc,
       aes(x=time,
           y=conc,
           shape=BLQ,
           group=subject)) +
  geom_ribbon(data=my_conc[my_conc$include_aucall,],
              aes(ymin=0, ymax=conc),
              fill="lightblue") +
  geom_line() +
  geom_point(size=4) +
  scale_x_continuous(breaks=my_conc$time) +
  theme(legend.position=c(0.8, 0.8))
```

# AUC to Infinity (AUC~$\infty$~)

AUC~0-$\infty$~ is commonly used for single-dose data.  It calculates the AUC~0-last~ and then extrapolates to $\infty$ using the estimated half-life.  Two starting points are used to estimate from `tlast` to $\infty$, the observed or half-life predicted concentration at `tlast` (`clast.obs` and `clast.pred`).

The two figures below illustrate the integration with AUC~0-$\infty$,obs~ and AUC~0-$\infty$,pred~.  The difference between the two figures is most evident at time=8 where there is a discontinuity in integration at `tlast` due to using `clast.pred` after that point and `clast.obs` before that point.  (To illustrate the integration differences, BLQ indicator shapes have been removed.  BLQ is handled identically to previous figures.)

```{r aucinf, fig.width=6}
# Add one row to illustrate extrapolation beyond observed data
my_conc <-
  rbind(my_conc,
        data.frame(conc=NA,
                   time=36,
                   subject=1,
                   BLQ=NA,
                   measured=FALSE,
                   include_auclast=FALSE,
                   include_aucall=FALSE))
# Extrapolate concentrations for aucinf.obs
my_conc$conc_aucinf.obs <- my_conc$conc
my_conc$conc_aucinf.obs[my_conc$BLQ | is.na(my_conc$BLQ)] <-
  interp.extrap.conc(conc=my_conc$conc,
                     time=my_conc$time,
                     time.out=my_conc$time[my_conc$BLQ | is.na(my_conc$BLQ)],
                     lambda.z=as.data.frame(results_obj)$PPORRES[as.data.frame(results_obj)$PPTESTCD %in% "lambda.z"])

# Extrapolate concentrations for aucinf.pred
my_conc$conc_aucinf.pred <- my_conc$conc
my_conc$conc_aucinf.pred[my_conc$BLQ | is.na(my_conc$BLQ)] <-
  interp.extrap.conc(conc=my_conc$conc,
                     time=my_conc$time,
                     time.out=my_conc$time[my_conc$BLQ | is.na(my_conc$BLQ)],
                     lambda.z=as.data.frame(results_obj)$PPORRES[as.data.frame(results_obj)$PPTESTCD %in% "lambda.z"],
                     clast=as.data.frame(results_obj)$PPORRES[as.data.frame(results_obj)$PPTESTCD %in% "clast.pred"])
my_conc$conc_aucinf.pred[my_conc$time == tlast] <-
  as.data.frame(results_obj)$PPORRES[as.data.frame(results_obj)$PPTESTCD %in% "clast.pred"]
```
```{r aucinf-visualization, eval=requireNamespace("ggplot2")}
ggplot(my_conc[!is.na(my_conc$conc),],
       aes(x=time,
           y=conc,
           #shape=BLQ,
           group=subject)) +
  geom_ribbon(data=my_conc,
              aes(ymin=0, ymax=conc_aucinf.obs),
              fill="lightblue") +
  geom_line() +
  #geom_point(size=2) +
  scale_x_continuous(breaks=my_conc$time) +
  theme(legend.position=c(0.8, 0.8)) +
  labs(title="Extrapolation using AUCinf,obs")

ggplot(my_conc[!is.na(my_conc$conc),],
       aes(x=time,
           y=conc,
           #shape=BLQ,
           group=subject)) +
  geom_ribbon(
    data=arrange(
      bind_rows(mutate(my_conc,
                       before=FALSE),
                mutate(filter(my_conc, time == tlast),
                       conc_aucinf.pred=conc,
                       before=TRUE)),
      time, desc(before)),
    aes(ymin=0,
        ymax=conc_aucinf.pred),
    fill="lightblue") +
  geom_line() +
  #geom_point(size=2) +
  scale_x_continuous(breaks=my_conc$time) +
  theme(legend.position=c(0.8, 0.8)) +
  labs(title="Extrapolation using AUCinf,pred")
```

# Partial AUCs

Partial AUCs integrate part of the area within a time range of interest.  Partial AUCs are often of interest to assess bioequivalence with more detail than AUC~0-$\infty$~ or AUC~0-last~ may indicate.  Within PKNCA, partial AUCs are calculated with the interval AUC (`aucint`) family of parameters:  `aucint.last`, `aucint.all`, `aucint.inf.obs`, and `aucint.inf.pred` integrate over the full interval from `start` to `end` while handling the region after `tlast` the same way as `auclast`, `aucall`, `aucinf.obs`, and `aucinf.pred`, respectively.

Request an `aucint` parameter like any other interval parameter with the `start` and `end` of the interval defining the range of integration.  When the interval boundaries do not match observed time points, concentrations at the boundaries are automatically interpolated (or extrapolated, according to the method of the specific parameter), so no manual data preparation is required.  In the example below, the concentration at the end time of 1.5 is not in the observed data, and it is interpolated automatically during the calculation.

When dosing data are given, the `aucint` parameters are dose-aware:  the profile being integrated ends at the first dose at or after the end of the interval, so a concentration measured after that dose is never interpolated back into the interval.  Only the start and end of the interval are estimated; a dose within the interval is integrated across using the concentrations measured on either side of it.  The method (`PPANMETH`) column reports whether the calculation was dose-aware and how the region after `tlast` was handled -- with the half-life, with `AUCall` (when `aucint.inf.obs` or `aucint.inf.pred` was requested but the half-life is not estimable), or not at all when the interval ends at or before `tlast`.

```{r partial_auc}
data_aucint_obj <-
  PKNCAdata(conc_obj,
            intervals=data.frame(start=0, end=c(2, 1.5), aucint.last=TRUE))
results_aucint_obj <- pk.nca(data_aucint_obj)
kable(as.data.frame(results_aucint_obj))
```

The area under the first moment curve (AUMC) has matching interval parameters (`aumcint.last`, `aumcint.all`, and the other `aumcint*` parameters), as do the intravenous C~0~ back-extrapolated AUMC parameters (`aumciv*`), in the development version of PKNCA.
