---
title: "Half-life calculation with Tobit regression"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Half-life calculation with Tobit regression}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r pre-setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(PKNCA)
```

# Half-life calculation with Tobit regression

Half-life calculation with Tobit regression allows inclusion of concentrations
that are below the lower limit of quantification in the half-life estimate.

## Comparison to semi-log regression

Typical half-life calculation uses curve-stripping semi-log regression of the
natural logarithm of the concentration by time.  Based on the use of the
logarithm of the concentration, concentrations below the lower limit of
quantification (LLOQ) which are set to zero are ignored.

Tobit regression allows inclusion of the concentrations below the LLOQ in the
half-life calculation.  Tobit regression for half-life is equivalent to using
Beal's M3 method in population pharmacokinetic (PK) models.

With Tobit regression, a line is fit using maximum likelihood.  For points above
the LLOQ, the likelihood is based on the probability density at the observed
concentration.  For points below the LLOQ, the likelihood is based on the
cumulative probability distribution function from negative infinity to the limit
of quantification.

## Automatic point selection

## Automatic point selection with semi-log regression

With semi-log regression, the typical method used to automatically select
concentrations for inclusion in the half-life estimate is to:

1. Omit all concentrations that are missing.
2. Omit all concentrations that are below the LLOQ.
3. Estimate the half-life for each set of points from the first concentration
   measure after T~max~ to the third measure before T~last~.
4. Select the best half-life with the following criteria, in order:
    1. The adjusted r-squared must be within a tolerance factor (typically
       0.0001) of the largest adjusted r-squared.
    2. The $\lambda_z$ value (slope for the half-life line) must be positive;
       in other words, the half-life slope must be decreasing.
    3. If multiple choices of points fit the above criteria, choose the one with
       the most concentration measurements included.

For comparison with PKNCA, note that Phoenix WinNonlin switches the order for
selection of 4.1 and 4.2 above.  So, if the best adjusted r-squared is for an
increasing slope but there is another adjusted r-squared with a decreasing
slope, Phoenix will report the half-life.

## Automatic point selection with Tobit regression

With Tobit regression, the method is generally similar to the semi-log
regression with two changes.  The first change is that concentrations below the
LLOQ are retained in the estimate.  The second change is that the adjusted
r-squared is not possible to calculate when including points below the LLOQ, so
the minimum standard deviation estimate is used.

The selection method below results in effectively the same estimates for
half-life when all points are above the LLOQ and improved estimates for
half-life when some points are below the LLOQ.  Future research may investigate
optimization of this method.

The steps for Tobit regression are:

1. Omit all concentrations that are missing.
2. Estimate the half-life for each set of points from the first concentration
   measure after T~max~ to the third measure before T~last~ while including all
   points below the LLOQ after T~last~.
3. Select the best half-life with the following criteria, in order:
    1. The estimated standard deviation of the slope is minimized.
    2. The $\lambda_z$ value (slope for the half-life line) must be positive;
       in other words, the half-life slope must be decreasing.

# Comparison of Tobit and semi-log regression

In almost all scenarios, Tobit regression using the algorithm above improves the
half-life estimate compared to semi-log regression.  In the figure below,
concentration-time profiles were simulated with 1-, 2-, and 3-compartment linear
PK models with intravenous or extravascular administration and a variety of
compartmental model parameters.  The true half-life was calculated based on the
compartmental model parameters.  Then, the ratio of the estimated to true
half-life was calculated.  Values closer to 1 indicate a better fit and values
farther from 1 indicate a poorer fit.

![The empirical cumulative distribution function for ratio of estimated to theoretical half-lives for Tobit regression (aqua line) and semi-log regression (red line) are shown.](v06-half-life-calculation-tobit_figure_1.svg)

Tobit regression performs universally better than least-squares up to the
estimated theoretical half-life, and better at >2-fold above the theoretical
half-life while least-squares performs slightly better between the theoretical
and 2-fold above.  The fact that the Tobit regression cumulative distribution
function is closer to 1 across the range of simulations indicates that Tobit
regression provides better half-life estimate across a broad range of data.

# Usage

## Basic usage: a single subject

The `hl_method` argument to `pk.calc.half.life()` selects the estimation
method.  When using `"tobit"`, the `lloq` argument is required and must be
either a scalar (one LLOQ applied to all observations) or a vector the same
length as `conc`.

### Example 1: All observations above the LLOQ

When no observations are below the LLOQ, Tobit and log-linear regression
produce essentially the same answer.

```{r example-no-blq}
conc_no_blq <- c(10, 5, 2.5, 1.25, 0.625)
time_points  <- c(0,  1, 2,   3,    4)
lloq         <- 0.1   # well below all observations

hl_loglinear <- pk.calc.half.life(
  conc = conc_no_blq,
  time = time_points,
  allow.tmax.in.half.life = TRUE
)

hl_tobit <- pk.calc.half.life(
  conc = conc_no_blq,
  time = time_points,
  lloq = lloq,
  hl_method = "tobit",
  allow.tmax.in.half.life = TRUE
)

cat("Log-linear half-life:", round(hl_loglinear$half.life, 3), "\n")
cat("Tobit half-life:     ", round(hl_tobit$half.life, 3),    "\n")
```

Note the columns that are returned:

```{r example-no-blq-cols}
# Log-linear returns r-squared and related diagnostics
names(hl_loglinear)

# Tobit returns tobit_residual and BLQ point counts instead
names(hl_tobit)
```

### Example 2: BLQ observations before T~last~

When some observations are below the LLOQ, the log-linear method drops them
entirely.  Tobit regression treats them as left-censored, which can provide a
better estimate.

```{r example-with-blq}
lloq <- 1.0
conc_true  <- c(10, 5, 2.5, 1.25, 0.5, 0.2)
time_points <- c(0,  1, 2,   3,    4,   5)
# Observations below the LLOQ are reported as zero
conc_obs <- ifelse(conc_true < lloq, 0, conc_true)

cat("Observed concentrations:", conc_obs, "\n")
cat("LLOQ:", lloq, "\n")
cat("Number of BLQ observations:", sum(conc_obs < lloq), "\n")

hl_loglinear <- pk.calc.half.life(
  conc = conc_obs,
  time = time_points,
  allow.tmax.in.half.life = TRUE
)

hl_tobit <- pk.calc.half.life(
  conc = conc_obs,
  time = time_points,
  lloq = lloq,
  hl_method = "tobit",
  allow.tmax.in.half.life = TRUE
)

cat("\nLog-linear (drops BLQ):\n")
cat("  Half-life:      ", round(hl_loglinear$half.life, 3), "\n")
cat("  Points used:    ", hl_loglinear$lambda.z.n.points, "\n")

cat("\nTobit (includes BLQ as censored):\n")
cat("  Half-life:      ", round(hl_tobit$half.life, 3), "\n")
cat("  Total points:   ", hl_tobit$lambda.z.n.points, "\n")
cat("  BLQ points:     ", hl_tobit$lambda.z.n.points_blq, "\n")
cat("  Tobit residual: ", round(hl_tobit$tobit_residual, 4), "\n")
```

## Using the global option

To use Tobit regression for all half-life calculations in a session, set the
global option.  Calculations that do not explicitly set `hl_method` will then
use Tobit regression by default.

```{r global-option, eval=FALSE}
PKNCA.options(hl_method = "tobit")

# Now all pk.calc.half.life() calls default to Tobit
# Reset to the package default when done:
PKNCA.options(default = TRUE)
```

## Full NCA workflow with PKNCAdata

The `lloq` argument can be passed through the interval calculation via the
`options` list when running a full NCA.

```{r full-nca, eval=FALSE}
# Suppose d_conc is your concentration data frame with columns
# subject, time, conc, and your LLOQ is 1.0

o_conc <- PKNCAconc(d_conc, conc ~ time | subject)
o_dose <- PKNCAdose(d_dose, dose ~ time | subject)
o_data <- PKNCAdata(
  o_conc, o_dose,
  intervals = data.frame(start = 0, end = Inf, half.life = TRUE),
  options = list(hl_method = "tobit")
)
# Note: lloq must be provided per-subject in the concentration data or
# as a scalar in the options when using the Tobit method in pk.nca().
```

## Controlling Tobit point selection

The `tobit_n_points_penalty` option controls how the number of points
influences window selection.  With the default value of 0 the window with the
smallest Tobit residual is selected.  Positive values penalize larger windows,
requiring them to have a meaningfully lower residual to be preferred.

```{r penalty-option, eval=FALSE}
# Penalise larger windows slightly
PKNCA.options(tobit_n_points_penalty = 0.5)
PKNCA.options(default = TRUE)  # reset
```

The `tobit_optim_control` option passes a list of control parameters directly
to `stats::optim()`.  This is rarely needed but can help when the default
optimization settings fail to converge for unusual concentration-time profiles.

```{r optim-control, eval=FALSE}
# Increase the maximum number of iterations
PKNCA.options(tobit_optim_control = list(maxit = 1000))
PKNCA.options(default = TRUE)  # reset
```
