---
title: "Half-Life Calculation"
author: "Bill Denney"
output:
  rmarkdown::html_vignette:
    toc: yes
    toc_depth: 6
vignette: >
  %\VignetteIndexEntry{Half-Life Calculation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, echo=FALSE, include=FALSE}
library(PKNCA)
library(dplyr)
```

# Introduction

Half-life is calculated by fitting the natural logarithm of concentration by time.  The default calculation method is curve stripping (described in more detail below).  Manual half-life points with no automated half-life selection can be performed, or specific points can be excluded while still performing curve stripping.  Half-life may also be estimated with Tobit regression, which includes concentrations below the lower limit of quantification as left-censored values; see the "Half-life calculation with Tobit regression" vignette for that method.

# Curve Stripping Method

When automatic point selection is performed for curve stripping, the algorithm described below is used.

## Select the Points

All sets of points that are applicable according to the current options are selected.

* Drop all BLQ values, then
* Drop all points at or before the end of the last dose administration, including infusion duration (if dosing information is provided), then
* Choose all sets of points that start from the $T_{last}$ and step back:
    * at least `r PKNCA.options("min.hl.points")` points (customizable with `PKNCA.options("min.hl.points")`)
    * `r c("Not including", "Including")[PKNCA.options("allow.tmax.in.half.life") + 1]` $T_{max}$ (customizable with `PKNCA.options("allow.tmax.in.half.life")`)

As a specific example, if measurements were at 0, 1, 2, 3, 4, 6, 8, 12, and 24 hours; if $T_{last}$ is 12 hours; and if $T_{max}$ is 1 hour then the default point sets that would be fit are:

1. 6, 8, and 12 hours;
2. 4, 6, 8, and 12 hours;
3. 3, 4, 6, 8, and 12 hours; and
4. 2, 3, 4, 6, 8, and 12 hours.

If `PKNCA.options("min.hl.points")` were set to `4`, then the 6, 8, and 12 hour set would not be fit.  If `PKNCA.options("allow.tmax.in.half.life")` were set to `TRUE`, then 1, 2, 3, 4, 6, 8, and 12 hours would be fit.

## Select the Best Fit

After fitting all points, the best fit among the set of possible fit is selected by the following rules:

1. $\lambda_z > 0$ and at the same time the selection statistic must be within a tolerance factor of the best value of that statistic.
    1. The statistic is the adjusted $r^2$ by default, and its tolerance is controlled by `PKNCA.options("adj.r.squared.factor")` which defaults to `r PKNCA.options("adj.r.squared.factor")`.
    2. Selecting on the unadjusted $r^2$ instead needs only `PKNCA.options("r.squared.factor")` set to the tolerance:  the two factors are alternatives, so setting either one takes the other out of use.  Because the unadjusted $r^2$ does not reward larger windows, it usually selects fewer points than the adjusted $r^2$ does.
    3. These rules must be met simultaneously, so if the best value of the statistic is for a line with $\lambda_z \leq 0$, the half-life may end up being unreportable.
2. If fitting the log-linear concentration-time line fails, then it is not the best line.
3. If more than one fit still meets the criteria above, then choose the fit with the most points included.

```{r check-ggplot, include=!requireNamespace("dplyr"), results="asis"}
cat("dplyr is required for this vignette to work correctly.  Please install the dplyr library and retry building the vignette.")
```

## Example

```{r normal-example}
# Perform calculations for subject 1, only
data_conc <- as.data.frame(datasets::Theoph)[datasets::Theoph$Subject == 1, ]

# Keep all points
conc_obj <-
  PKNCAconc(
    data_conc,
    conc~Time|Subject
  )

# Only calculate half-life and parameters required for half-life
current_intervals <- data.frame(start=0, end=Inf, half.life=TRUE)
data_obj <- PKNCAdata(conc_obj, intervals=current_intervals)
result_obj <- pk.nca(data_obj)

# Extract the results for subject 1 
as.data.frame(result_obj)
```


# Manual Point Selection

Curve stripping chooses the points automatically, and it is what happens when
nothing is specified.  Two optional columns given to `PKNCAconc()` change that:
`exclude_half.life` drops specific points and then curve strips whatever
remains, while `include_half.life` names the exact points to fit and switches
curve stripping off entirely.  The decision tree below shows how to choose
between them and what each one changes.

## Choosing a Point Selection Method

```{r decision-tree-options, echo=FALSE}
# Named in both the figure and the text below.  Fail the build if one is
# renamed or removed rather than describing options that no longer exist.
hl_auto_options <-
  c("min.hl.points", "allow.tmax.in.half.life", "adj.r.squared.factor",
    "r.squared.factor")
stopifnot(all(hl_auto_options %in% names(PKNCA.options())))
hl_auto_options_md <-
  paste0(
    paste0("`", utils::head(hl_auto_options, -1), "`", collapse = ", "),
    ", and `", utils::tail(hl_auto_options, 1), "`"
  )
```

```{r decision-tree, echo=FALSE, fig.width=8.4, fig.height=5.2, dpi=192, out.width="100%", fig.alt="Decision tree for choosing between automatic curve stripping, exclude_half.life, and include_half.life.  Automatic curve stripping is the default and needs neither column.  exclude_half.life drops the points flagged TRUE and curve strips the rest.  include_half.life keeps only the points flagged TRUE and fits them directly.  In every case, BLQ and zero concentrations and points at or before the end of the last dose are dropped before fitting."}
# The tree is drawn in the coordinate system of the figure itself (840 wide by
# 520 tall, y increasing downward) so that positions read as pixels.
tree_box <- function(x, y, w, h, fill = "white") {
  rect(x, y, x + w, y + h, border = "black", lwd = 1.5, col = fill)
}
tree_arrow <- function(x, y0, y1) {
  arrows(x, y0, x, y1, length = 0.08, angle = 20, lwd = 1.5)
}

cex_body <- 0.78
cex_head <- 0.9
branch_x <- c(140, 420, 700)
branch_x_left <- branch_x - 125

op <- par(mar = rep(0, 4))
plot.new()
plot.window(xlim = c(0, 840), ylim = c(520, 0), xaxs = "i", yaxs = "i")

tree_box(45, 15, 750, 45)
text(420, 38, "How should the points for the half-life fit be chosen?",
     font = 2, cex = cex_head)
segments(420, 60, 420, 92, lwd = 1.5)
segments(140, 92, 700, 92, lwd = 1.5)
for (x in branch_x) tree_arrow(x, 92, 120)

for (x in branch_x_left) tree_box(x, 120, 250, 60)
text(
  branch_x, 150,
  c(
    "Let PKNCA choose them\n(the default)",
    "Curve strip, but drop\nspecific known-bad points",
    "Use exactly the points I\nname, with no curve stripping"
  ),
  cex = cex_body
)
for (x in branch_x) tree_arrow(x, 180, 215)

for (x in branch_x_left) tree_box(x, 215, 250, 60)
text(140, 237, "Set neither column", cex = cex_body)
text(c(420, 700), 237, c("exclude_half.life", "include_half.life"),
     cex = cex_body, family = "mono")
text(branch_x, 257,
     c("on PKNCAconc()", "TRUE drops that point", "TRUE keeps that point"),
     cex = cex_body)
for (x in branch_x) tree_arrow(x, 275, 310)

tree_box(15, 310, 810, 70, fill = "grey95")
text(420, 331, "In every case, before any line is fit:", font = 2, cex = cex_body)
text(
  420, 358,
  paste(
    "drop BLQ and zero concentrations, and drop points at or before the end of the",
    "last dose (including infusion duration)",
    sep = "\n"
  ),
  cex = cex_body
)
for (x in branch_x) tree_arrow(x, 380, 405)

tree_box(15, 405, 530, 100)
text(280, 428, "Automatic point selection", font = 2, cex = cex_body)
text(280, 455, "Every allowed set of points is fit and the best is kept, using",
     cex = cex_body)
text(280, 477,
     paste(
       paste(utils::head(hl_auto_options, 2), collapse = ", "),
       paste(utils::tail(hl_auto_options, -2), collapse = ", "),
       sep = ",\n"
     ),
     cex = cex_body, family = "mono")

tree_box(575, 405, 250, 100)
text(700, 428, "Manual point selection", font = 2, cex = cex_body)
text(700, 466,
     sprintf(
       "One fit through exactly those\npoints.  The %d options at\nleft are ignored.",
       length(hl_auto_options)
     ),
     cex = cex_body)
par(op)
```

Two parts of the tree are easy to overlook:

* Only `include_half.life` turns off automatic selection.  `exclude_half.life`
  removes the flagged points and then runs the usual curve stripping on the
  rest, so `r hl_auto_options_md` still apply to the points that remain.
* The middle band of the tree applies to `include_half.life` as well.  Naming a
  point does not force it into the fit; BLQ and zero concentrations, and points
  at or before the end of the last dose, are dropped first either way.
  Including eight points that span a four-hour infusion therefore yields a
  five-point fit, not an eight-point fit.

## Column Requirements

Both columns must be logical (`TRUE`/`FALSE`/`NA`); a non-logical column (for
example, character `"yes"`) is an error.  Whether a column applies is decided
separately for each interval, using the rules below.

| Situation for an interval | Result |
| ------------------------- | ------ |
| The column was not given to `PKNCAconc()` | The method is not used |
| Every value is `NA` | The method is not used; the interval is treated as though the column had not been given |
| Some values are `NA` | The `NA` values are treated as `FALSE` |
| Any value is not `NA`, even if every value is `FALSE` | The column counts as in use |
| Both columns are in use | Error: "Cannot both include and exclude half-life points for the same interval" |

Because an all-`FALSE` column still counts as in use, initialize these columns
to `NA` (rather than `FALSE`) wherever the corresponding method should not
apply.

## Exclusion of Specific Points with Curve Stripping

In some cases, specific points will be known outliers, or there may be another reason to exclude specific points.  And, with those points excluded, the half-life should be calculated using the normal curve stripping methods described above.

To exclude specific points but otherwise use curve stripping, use the `exclude_half.life` option as the column name in the concentration dataset for `PKNCAconc()` as illustrated below.

```{r exclude-points}
data_conc$exclude_hl <- data_conc$Time == 12.12
# Confirm that we will be excluding exactly one point
stopifnot(sum(data_conc$exclude_hl) == 1)

# Drop one point
conc_obj_exclude1 <-
  PKNCAconc(
    data_conc,
    conc~Time|Subject,
    exclude_half.life="exclude_hl"
  )

data_obj_exclude1 <- PKNCAdata(conc_obj_exclude1, intervals=current_intervals)

# Perform the calculations
result_obj_exclude1 <- pk.nca(data_obj_exclude1)

# Results differ when excluding the 12-hour point for subject 1 (compare to
# example in the previous section)
as.data.frame(result_obj_exclude1)
```

## Specification of the Exact Points for Analysis

In other cases, the exact points to use for half-life calculation are known, and automatic point selection with curve stripping should not be performed.

To use only specific points and bypass automatic curve stripping, use the `include_half.life` option as the column name in the concentration dataset for `PKNCAconc()` as illustrated below.

```{r include-points}
data_conc$include_hl <- data_conc$Time > 3
# Confirm that we will be including exactly six points
stopifnot(sum(data_conc$include_hl) == 6)

# Use only these points for the half-life
conc_obj_include6 <-
  PKNCAconc(
    data_conc,
    conc~Time|Subject,
    include_half.life="include_hl"
  )

data_obj_include6 <- PKNCAdata(conc_obj_include6, intervals=current_intervals)

# Perform the calculations
result_obj_include6 <- pk.nca(data_obj_include6)

# Results differ when including 6 points (compare to example in the previous
# section)
as.data.frame(result_obj_include6)
```
