---
title: "Unit Assignment and Conversion with PKNCA"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Unit Assignment and Conversion with PKNCA}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
suppressPackageStartupMessages(
  library(PKNCA)
)
```

# Introduction

PKNCA can assign and convert units for reporting. There are two ways to provide
units to PKNCA: via the `units` argument to `PKNCAdata()` or by specifying units
with `PKNCAconc()` and/or `PKNCAdose()`. If you provide the units argument to
`PKNCAdata()`, units given to `PKNCAconc()` or `PKNCAdose()` are ignored.

# Examples of each way to add units

## Steps to add units to an NCA analysis from the data

For more details on parts of this NCA calculation example unrelated to units, see the [theophylline example vignette](v02-example-theophylline.html).

Provide the units for concentration (`concu`), time (`timeu`), and amount
(`amountu`) to the `PKNCAconc()` function and for dose (`doseu`) to the
`PKNCAdose()` function.

```{r create-conc-dose-obj-auto}
d_conc <- as.data.frame(datasets::Theoph)
d_conc$concu_col <- "mg/L"
d_conc$timeu_col <- "hr"
d_dose <- datasets::Theoph[datasets::Theoph$Time == 0, c("Dose", "Time", "Subject")]
d_dose$doseu_col <- "mg/kg"
o_conc <- PKNCAconc(d_conc, conc~Time|Subject, concu = "concu_col", timeu = "timeu_col")
o_dose <- PKNCAdose(d_dose, Dose~Time|Subject, doseu = "doseu_col")
```

Then, create the data object the same way as typical. Results will have units.

```{r create-data-obj-auto}
o_data <- PKNCAdata(o_conc, o_dose)
o_nca <- pk.nca(o_data)
summary(o_nca)
```

When units are provided through `PKNCAconc()` and `PKNCAdose()` like this,
`PKNCAdata()` builds the units table automatically during construction by
calling `pknca_units_table()` on the `PKNCAdata` object.  You can also call
`pknca_units_table()` on a `PKNCAdata` object yourself to build or inspect that
units table directly.

It is also possible to specify the units without them coming from columns in the
data.

```{r create-auto-direct}
d_conc <- as.data.frame(datasets::Theoph)
d_dose <- datasets::Theoph[datasets::Theoph$Time == 0, c("Dose", "Time", "Subject")]
o_conc <- PKNCAconc(d_conc, conc~Time|Subject, concu = "mg/L", timeu = "hr")
o_dose <- PKNCAdose(d_dose, Dose~Time|Subject, doseu = "mg/kg")
o_data <- PKNCAdata(o_conc, o_dose)
o_nca <- pk.nca(o_data)
summary(o_nca)
```

And, you can perform automatic unit conversions as long as the unit conversions
are defined without more information (e.g. convert between mass or time units).
For more complex conversions, see the information below.

```{r create-auto-direct-pref}
d_conc <- as.data.frame(datasets::Theoph)
d_dose <- datasets::Theoph[datasets::Theoph$Time == 0, c("Dose", "Time", "Subject")]
o_conc <- PKNCAconc(d_conc, conc~Time|Subject, concu = "mg/L", timeu = "hr", concu_pref = "ug/L", timeu_pref = "day")
o_dose <- PKNCAdose(d_dose, Dose~Time|Subject, doseu = "mg/kg")
o_data <- PKNCAdata(o_conc, o_dose)
o_nca <- pk.nca(o_data)
summary(o_nca)
```

## Steps to manually add units to an NCA analysis

For more details on parts of this NCA calculation example unrelated to units, see the [theophylline example vignette](v02-example-theophylline.html).

```{r create-conc-dose-obj}
o_conc <- PKNCAconc(as.data.frame(datasets::Theoph), conc~Time|Subject)
d_dose <- datasets::Theoph[datasets::Theoph$Time == 0, c("Dose", "Time", "Subject")]
o_dose <- PKNCAdose(d_dose, Dose~Time|Subject)
```

The difference from a calculation without units comes when setting up the `PKNCAdata` object.  You will add the units with the `units` argument.  

Since no urine or other similar collection is performed, the `amountu` argument is omitted for `pknca_units_table()`.

```{r create-data-obj}
d_units <-
  pknca_units_table(
    concu="mg/L", doseu="mg/kg", timeu="hr",
    # use molar units for concentrations and AUCs
    conversions=
      data.frame(
        PPORRESU=c("(mg/kg)/(hr*mg/L)", "(mg/kg)/(mg/L)", "mg/L", "hr*mg/L"),
        PPSTRESU=c("L/hr/kg", "L/kg", "mmol/L", "hr*mmol/L"),
        conversion_factor=c(NA, NA, 1/180.164, 1/180.164)
      )
  )

o_data <- PKNCAdata(o_conc, o_dose, units=d_units)
o_nca <- pk.nca(o_data)
summary(o_nca)
```

# Prepare a Unit Assignment and Conversion Table

A unit assignment and conversion table can be generated as a data.frame to use
with the `pknca_units_table()` function or manually.

The simplest method takes each of the types of units for inputs and
automatically generates the units for each NCA parameter.

```{r create-units-auto}
d_units_auto <- pknca_units_table(concu="ng/mL", doseu="mg", amountu="mg", timeu="hr")
# Show a selection of the units generated
d_units_auto[d_units_auto$PPTESTCD %in% c("cmax", "tmax", "auclast", "cl.obs", "vd.obs"), ]
```

As you see above, the default units table has a column for the `PPTESTCD`
indicating the parameter.  And, the column `PPORRESU` indicates what the default
units are.

Without unit conversion, the units for some parameters (notably clearances and
volumes) are not so useful.  You can add a conversion table to make any units
into the desired units.  For automatic conversion to work, the units must always
be convertible (by the `units` library).  Notably for automatic conversion, you
cannot go from mass to molar units since there is not a unique conversion from
mass to moles.

Loading the PKNCA package adds a unit of `"fraction"` so that it is usable for
fraction excreted (`fe`).

```{r create-units-semi-manual}
d_units_clean <-
  pknca_units_table(
    concu="ng/mL", doseu="mg", amountu="ng", timeu="hr",
    conversions=
      data.frame(
        PPORRESU=c("mg/(hr*ng/mL)", "mg/(ng/mL)", "hr", "ng/mg"),
        PPSTRESU=c("L/hr", "L", "day", "fraction")
      )
  )
# Show a selection of the units generated
d_units_clean[d_units_clean$PPTESTCD %in% c("cmax", "tmax", "auclast", "cl.obs", "vd.obs", "fe"), ]
```

Now, the units are much cleaner to look at.

To do a conversion that is not possible directly with the `units` library, you can add the conversion factor manually by adding the `conversion_factor` column.  You can mix-and-match manual and automatic modification by setting the `conversion_factor` column to `NA` when you want automatic conversion.  In the example below, we convert concentration units to molar.  Note that AUC units are not set to molar because we did not specify that conversion; all conversions must be specified.

```{r create-units-manual}
d_units_clean_manual <-
  pknca_units_table(
    concu="ng/mL", doseu="mg", amountu="mg", timeu="hr",
    conversions=
      data.frame(
        PPORRESU=c("mg/(hr*ng/mL)", "mg/(ng/mL)", "hr", "ng/mL"),
        PPSTRESU=c("L/hr", "L", "day", "nmol/L"),
        conversion_factor=c(NA, NA, NA, 1000/123)
      )
  )
# Show a selection of the units generated
d_units_clean_manual[d_units_clean_manual$PPTESTCD %in% c("cmax", "tmax", "auclast", "cl.obs", "vd.obs"), ]
```

# What happens when units are missing for some parameters?

A hand-made or hand-edited units table may not have a row for every parameter
that an analysis requests.  By default, `pk.nca()` raises an error when units
are provided for some but not all requested parameters so that results are not
unintentionally reported without units.  Setting the
`allow_partial_missing_units` option to `TRUE` converts that error to a warning
naming the parameters without units, and those parameters are reported without
units.  The option can be set for a single analysis with the `options` argument
to `PKNCAdata()` (as below) or globally with
`PKNCA.options(allow_partial_missing_units = TRUE)`.

```{r partial-missing-units}
o_conc <- PKNCAconc(as.data.frame(datasets::Theoph), conc~Time|Subject)
d_dose <- datasets::Theoph[datasets::Theoph$Time == 0, c("Dose", "Time", "Subject")]
o_dose <- PKNCAdose(d_dose, Dose~Time|Subject)
d_units_full <- pknca_units_table(concu="mg/L", doseu="mg/kg", timeu="hr")
# Drop the tmax row to simulate an incomplete units table
d_units_partial <- d_units_full[!d_units_full$PPTESTCD %in% "tmax", ]
o_data_partial <-
  PKNCAdata(
    o_conc, o_dose,
    units=d_units_partial,
    options=list(allow_partial_missing_units=TRUE)
  )
o_nca_partial <- pk.nca(o_data_partial)
summary(o_nca_partial)
```

# How do I add different unit conversions for different analytes?

Sometimes, when multiple analytes are used and, for example, molar outputs are desired while inputs are in mass units.  Different unit conversions may be required for different inputs.

Different unit conversions can be used by adding the grouping column to the units specification.

Start by setting up a concentration dataset with two analytes.  Since the dosing doesn't have an "Analyte" column, it will be matched to all concentration measures for the subject.

```{r different-units-conc-dose-setup}
d_conc_theoph <- as.data.frame(datasets::Theoph)
d_conc_theoph$Analyte <- "Theophylline"
# Approximately 6% of theophylline is metabolized to caffeine
# (https://www.pharmgkb.org/pathway/PA165958541).  Let's pretend that means it
# has 6% of the theophylline concentration at all times.
d_conc_caffeine <- as.data.frame(datasets::Theoph)
d_conc_caffeine$conc <- 0.06*d_conc_caffeine$conc
d_conc_caffeine$Analyte <- "Caffeine"
d_conc <- rbind(d_conc_theoph, d_conc_caffeine)

d_dose <- unique(datasets::Theoph[datasets::Theoph$Time == 0,
                                  c("Dose", "Time", "Subject")])
```

Setup the units with an "Analyte" column to separate the units used.

```{r different-units-units-setup}
d_units_theoph <-
  pknca_units_table(
    concu="mg/L", doseu="mg/kg", timeu="hr",
    # use molar units for concentrations and AUCs
    conversions=
      data.frame(
        PPORRESU=c("(mg/kg)/(hr*mg/L)", "(mg/kg)/(mg/L)", "mg/L", "hr*mg/L"),
        PPSTRESU=c("L/hr/kg", "L/kg", "mmol/L", "hr*mmol/L"),
        conversion_factor=c(NA, NA, 1/180.164, 1/180.164)
      )
  )
d_units_theoph$Analyte <- "Theophylline"
d_units_caffeine <-
  pknca_units_table(
    concu="mg/L", doseu="mg/kg", timeu="hr",
    # use molar units for concentrations and AUCs
    conversions=
      data.frame(
        PPORRESU=c("(mg/kg)/(hr*mg/L)", "(mg/kg)/(mg/L)", "mg/L", "hr*mg/L"),
        PPSTRESU=c("L/hr/kg", "L/kg", "mmol/L", "hr*mmol/L"),
        conversion_factor=c(NA, NA, 1/194.19, 1/194.19)
      )
  )
d_units_caffeine$Analyte <- "Caffeine"
d_units <- rbind(d_units_theoph, d_units_caffeine)
```

Now, calculate adding the different units per analyte to the data object.

```{r different-units-calc}
o_conc <- PKNCAconc(d_conc, conc~Time|Subject/Analyte)
o_dose <- PKNCAdose(d_dose, Dose~Time|Subject)
o_data <- PKNCAdata(o_conc, o_dose, units=d_units)
o_nca <- pk.nca(o_data)
summary(o_nca)
```

# Date and time (POSIXct) input

Concentration and dose times may be given as date-times (POSIXct) instead of
numbers, as they often are in clinical data (for example, the SDTM `--DTC`
variables after parsing).  Date-times are measured in seconds, so `timeu` must
be `"s"` or omitted; set `timeu_pref` to the unit for calculations and
reporting.  `PKNCAdata()` converts the times to numbers relative to the first
dose in each group:  the groups are the grouping variables shared by the
concentration and dose formulas, so with `Subject` as the group, time 0 is each
subject's first dose.  For studies with several parts (or crossover periods),
add the part (or period) to both formulas, and each subject's first dose in each
part is time 0.

```{r datetime-input}
d_conc <- as.data.frame(datasets::Theoph)
d_dose <- datasets::Theoph[datasets::Theoph$Time == 0, c("Dose", "Time", "Subject")]
# Each subject was dosed at 08:00 on a different day
first_dose <- as.POSIXct("2024-01-15 08:00", tz = "UTC") + (as.numeric(as.character(d_dose$Subject)) - 1) * 86400
names(first_dose) <- as.character(d_dose$Subject)
d_conc$datetime <- first_dose[as.character(d_conc$Subject)] + d_conc$Time * 3600
d_dose$datetime <- first_dose[as.character(d_dose$Subject)]
o_conc <- PKNCAconc(d_conc, conc~datetime|Subject, concu = "mg/L", timeu_pref = "hr")
o_dose <- PKNCAdose(d_dose, Dose~datetime|Subject, doseu = "mg/kg")
o_data <- PKNCAdata(o_conc, o_dose)
# The time reference for each subject
head(o_data$time_reference)
# The concentration times are now hours after the first dose
head(o_data$conc$data[, c("Subject", "datetime", "conc")])
```

Automatic intervals are chosen as for numeric times, and manually specified
intervals are numeric times relative to the time reference, in the preferred
time unit.  The default single-dose intervals (the `single.dose.aucs` option,
0 to 24 and 0 to infinity) are written for hours, so with a preferred time unit
such as `"day"` or `"min"` they would end at 24 days or 24 minutes; PKNCA warns
when that happens, and you should give `intervals` or set `single.dose.aucs`
for that unit.  Results formatted for CDISC include the time reference as the
PPRFTDTC column.

```{r datetime-results}
o_nca <- pk.nca(o_data)
d_cdisc <- as.data.frame(o_nca, out_format = "cdisc")
head(d_cdisc[, c("Subject", "PPTESTCD", "PPORRES", "PPORRESU", "PPRFTDTC")])
```

A few rules keep the conversion unambiguous:

* Concentration and dose times must both be date-times; mixing numeric and
  date-time times is an error.
* All date-times must have the same time zone.  Differences are elapsed time,
  so a daylight saving time change within a named time zone (like
  `"America/New_York"`) is handled correctly.
* Dates (Date) are taken as midnight at the start of the date, with a warning.
* The dose formula must include the subject (as with `Subject` above), so that
  each subject has its own reference.  Sparse data are the exception:  every
  subject in a sparse group shares the group's dosing, so the reference is per
  group.
* Excluded doses are not used as the time reference.  A subject without an
  included dose time uses its first (not excluded) concentration instead, with
  a warning, and without dosing data every subject uses its first
  concentration.  The `time_reference_type` column of `time_reference` says
  which kind of reference each group has (`"first_dose"` or `"first_conc"`).
* Nominal times (`time.nominal`) are not converted.
