---
title: "Secondary Parameters (Renal Clearance, Bioavailability, and Ratios)"
output:
  rmarkdown::html_vignette:
    toc: yes
    toc_depth: 6
vignette: >
  %\VignetteIndexEntry{Secondary Parameters (Renal Clearance, Bioavailability, and Ratios)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r libraries}
library(PKNCA)
library(knitr)
```

```{r display-helper}
# check.interval.specification() fills in every registered parameter as FALSE,
# so an interval specification that has been through it has a few hundred
# columns.  Show only the ones saying something:  the times, the groups, the
# linkage, and the parameters actually requested.
kable_intervals <- function(intervals) {
  says_something <-
    vapply(
      X = intervals,
      FUN = function(x) !is.logical(x) || any(x, na.rm = TRUE),
      FUN.VALUE = TRUE
    )
  kable(intervals[, says_something, drop = FALSE])
}
```

# What a secondary parameter is

Almost every NCA parameter is calculated from the concentrations and doses
within one interval of one profile.  A few cannot be:

* Renal clearance divides the amount excreted into a urine collection by the
  plasma AUC over the same times.  The amount and the AUC are measured on
  different profiles.
* Bioavailability compares the dose-normalized exposure of one administration
  with a reference administration.
* A ratio (an accumulation ratio, a metabolite ratio) compares a parameter with
  the same parameter somewhere else.

PKNCA calls these **secondary parameters** and calculates them by linking two
intervals together.  `pknca_parameter_table()` says which parameters are
secondary:

```{r secondary-parameter-list}
d_params <- pknca_parameter_table()
kable(
  d_params[d_params$secondary, c("parameter", "concept", "sample_type")],
  row.names = FALSE
)
```

Two words are used throughout:

* The **requesting interval** is the intervals row that asks for the parameter.
  The result is reported there:  on that row's group, that row's times, and in
  that row's units.
* The **reference interval** is the row that supplies the value from the other
  profile.

Secondary parameters are never selected automatically by
`pknca_interval_table()` or the default interval specification, because one
interval cannot say what the other profile is.  They are requested by name, like
any other parameter.

# Linking two intervals by hand

Two columns of the interval specification make the link:

| Column | Meaning |
|---|---|
| `interval_id` | A name for this row, so that another row can point at it.  `NA` on rows nothing points at. |
| `<parameter>_ref` | The `interval_id` of the interval supplying the cross-profile value for `<parameter>`. |

The identifiers may be character names, factors, or numbers such as row indices.
Whatever they are, `interval_id` and the pointer columns must be comparable with
each other, so that a pointer can be matched to an identifier.

## Renal clearance, start to finish

These data have one subject with a plasma profile and a urine collection.  For
the urine rows, `conc` is the concentration measured in the collection and `vol`
is the collection volume:

```{r renal-data}
d_conc <- data.frame(
  subject = 1,
  PCSPEC = rep(c("plasma", "urine"), times = c(3, 2)),
  time = c(0, 12, 24, 0, 12),
  conc = c(10, 6, 2, 2, 1),
  vol  = c(NA, NA, NA, 100, 150)
)
o_conc <- PKNCAconc(d_conc, conc~time|PCSPEC+subject, volume = "vol")
```

The plasma row is given an `interval_id`, and the urine row requests `clr.last`
pointing at it:

```{r renal-intervals}
intervals_linked <-
  data.frame(
    PCSPEC = c("plasma", "urine"),
    start = 0, end = 24,
    interval_id = c("plasma024", NA),
    auclast = c(TRUE, FALSE),
    ae = c(FALSE, TRUE),
    clr.last = c(FALSE, TRUE),
    clr.last_ref = c(NA, "plasma024")
  )
kable(intervals_linked)
```

```{r renal-calculate}
o_data <-
  PKNCAdata(
    o_conc, intervals = intervals_linked,
    options = list(auc.method = "linear")
  )
results_linked <- pk.nca(o_data)
kable(as.data.frame(results_linked))
```

Every number can be checked by hand.  With linear trapezoids the plasma AUC is
$(10+6)/2 \times 12 + (6+2)/2 \times 12 = 144$, the amount excreted is
$2 \times 100 + 1 \times 150 = 350$, and the renal clearance is
$350/144 = `r signif(350/144, 4)`$.

Note where the result is reported:  on the urine row, which requested it, and
not on the plasma row that supplied the AUC.

## The reference interval gains what the link needs

`auclast` was requested explicitly above to make the arithmetic visible, but it
did not have to be.  A reference interval gains whatever the link reads from it,
the same way any parameter's dependencies are calculated, and without
announcement:

```{r renal-silent}
intervals_implicit <- intervals_linked
intervals_implicit$auclast <- c(FALSE, FALSE)
intervals_implicit$cmax <- c(TRUE, FALSE)
o_data_implicit <-
  PKNCAdata(
    o_conc, intervals = intervals_implicit,
    options = list(auc.method = "linear")
  )
results_implicit <- pk.nca(o_data_implicit)
kable(as.data.frame(results_implicit))
```

The `auclast` row is in the results because it was calculated, but it was not
one of the parameters asked for.  `filter_requested = TRUE` reports only the
requests:

```{r renal-filter-requested}
d_requested <- as.data.frame(results_implicit, filter_requested = TRUE)
kable(d_requested[, c("PCSPEC", "subject", "start", "end", "PPTESTCD", "PPORRES")])
```

# Letting PKNCA find the reference interval

Naming the reference by hand is not usually necessary.  When a secondary
parameter is requested with no pointer and no way to calculate it within its own
interval, PKNCA derives the reference from the data.  Here the specification
asks only about the urine collection:

```{r auto-intervals}
intervals_urine <-
  data.frame(PCSPEC = "urine", start = 0, end = 24, ae = TRUE, clr.last = TRUE)
o_data_auto <-
  PKNCAdata(
    o_conc, intervals = intervals_urine,
    options = list(auc.method = "linear")
  )
results_auto <- pk.nca(o_data_auto)
kable(as.data.frame(results_auto))
```

The message reports the interval that was derived, and the answer is the one the
hand-written linkage gives.  The derivation is mechanical rather than a list of
special cases:  the parameter is measured over an interval collection while
everything it references is a spot sample (this is what renal clearance is), the
concentration data declare a collection `volume`, and so the reference is looked
for among the profiles that have no collection volume.  Of those, each interval
takes the one differing from its own group in the fewest columns.

Derived intervals are used for the calculation only.  The intervals that come
back with the results are the ones that went in:

```{r auto-ephemeral}
identical(results_auto$data$intervals, check.interval.specification(intervals_urine))
```

## When more than one profile could be the reference

Adding a second spot-sample profile leaves no single nearest reference.  Rather
than guessing or stopping the analysis, PKNCA reports `NA` with the reason:

```{r auto-ambiguous}
d_conc_serum <-
  rbind(
    d_conc,
    data.frame(
      subject = 1, PCSPEC = "serum", time = c(0, 12, 24), conc = c(9, 5, 2),
      vol = NA_real_
    )
  )
o_conc_serum <-
  PKNCAconc(d_conc_serum, conc~time|PCSPEC+subject, volume = "vol")
intervals_three <-
  data.frame(
    PCSPEC = c("urine", "plasma", "serum"),
    start = 0, end = 24,
    ae = c(TRUE, FALSE, FALSE),
    auclast = c(FALSE, TRUE, TRUE),
    clr.last = c(TRUE, FALSE, FALSE)
  )
o_data_ambiguous <-
  PKNCAdata(
    o_conc_serum, intervals = intervals_three,
    options = list(auc.method = "linear")
  )
d_ambiguous <- as.data.frame(pk.nca(o_data_ambiguous))
kable(d_ambiguous[d_ambiguous$PPTESTCD %in% "clr.last", ])
```

Everything else in the analysis is still calculated.  One warning is raised per
parameter, however many intervals it affected, and the reason for each result is
in its `exclude` column.

## Steering the choice with `group_ref`

`PKNCAdata(group_ref = )` says which profiles may be references.  It settles the
ambiguity above:

```{r group-ref}
o_data_steered <-
  PKNCAdata(
    o_conc_serum, intervals = intervals_three,
    options = list(auc.method = "linear"),
    group_ref = data.frame(PCSPEC = "plasma")
  )
d_steered <- as.data.frame(pk.nca(o_data_steered))
kable(d_steered[d_steered$PPTESTCD %in% "clr.last", ])
```

`group_ref` also directs comparisons that the data cannot tell apart at all.
Nothing in a data set says which analyte is the parent and which the metabolite,
so a metabolite ratio has to be told:

```{r group-ref-parameter-specific}
d_conc_two <-
  rbind(
    data.frame(
      Analyte = rep(c("parent", "metabolite"), each = 5),
      PCSPEC = "plasma", subject = 1,
      time = rep(c(0, 1, 2, 4, 8), 2),
      conc = c(0, 10, 8, 5, 2, 0, 4, 3.5, 2, 0.8),
      vol = NA_real_, dur = 0
    ),
    data.frame(
      Analyte = rep(c("parent", "metabolite"), each = 2),
      PCSPEC = "urine", subject = 1,
      time = rep(c(0, 12), 2),
      conc = c(2, 1, 1, 0.5),
      vol = rep(c(100, 150), 2), dur = 12
    )
  )
o_conc_two <-
  PKNCAconc(
    d_conc_two, conc~time|Analyte+PCSPEC+subject,
    volume = "vol", duration = "dur"
  )
intervals_two <-
  data.frame(
    Analyte = c("parent", "metabolite", "parent", "metabolite"),
    PCSPEC = c("urine", "urine", "plasma", "plasma"),
    start = 0,
    end = c(24, 24, Inf, Inf),
    ae = c(TRUE, TRUE, FALSE, FALSE),
    clr.last = c(TRUE, TRUE, FALSE, FALSE),
    aucinf.obs = c(FALSE, FALSE, TRUE, FALSE),
    ratio.aucinf.obs = c(FALSE, FALSE, FALSE, TRUE)
  )
```

Renal clearance and the metabolite ratio need to be steered by different
columns:  the renal clearance of each analyte belongs with the plasma profile of
the same analyte, while the ratio compares one analyte with the other.  A
`parameter` column says which rows apply to which parameter, and the columns a
parameter's row leaves `NA` do not constrain it:

```{r group-ref-table}
group_ref_by_param <-
  data.frame(
    parameter = c("clr.last", "ratio.aucinf.obs"),
    PCSPEC = c("plasma", NA),
    Analyte = c(NA, "parent")
  )
kable(group_ref_by_param)
```

```{r group-ref-two-contrasts}
o_data_two <-
  PKNCAdata(
    o_conc_two, intervals = intervals_two,
    options = list(auc.method = "linear"),
    group_ref = group_ref_by_param
  )
d_two <- as.data.frame(pk.nca(o_data_two))
kable(d_two[d_two$PPTESTCD %in% c("clr.last", "ratio.aucinf.obs"), ])
```

A named list of data.frames, one per parameter, says the same thing:

```{r group-ref-list, eval = FALSE}
group_ref = list(
  clr.last = data.frame(PCSPEC = "plasma"),
  ratio.aucinf.obs = data.frame(Analyte = "parent")
)
```

A plain data.frame with no `parameter` column applies to every secondary
parameter.

# Writing the linkage into the intervals

`interval_add_secondary()` writes the same linkage the engine derives, but
visibly, into the intervals it returns.  Use it to see the linkage, to edit it,
or to link something the derivation cannot reach.

```{r add-secondary}
intervals_plain <-
  data.frame(
    PCSPEC = c("plasma", "urine"),
    start = 0, end = 24,
    auclast = c(TRUE, FALSE),
    ae = c(FALSE, TRUE)
  )
kable_intervals(
  interval_add_secondary(
    intervals_plain,
    param = "clr.last",
    reference = data.frame(PCSPEC = "plasma")
  )
)
```

The plasma row was given an identifier, the urine row was given the request and
the pointer, and nothing else changed.

`reference` is a data.frame describing the reference interval:  each of its
columns must match (and), and any of its rows may match (or).  When nothing
matches, the reference interval is created and the creation is reported:

```{r add-secondary-create}
intervals_urine_only <-
  data.frame(PCSPEC = "urine", start = 0, end = 24, ae = TRUE)
kable_intervals(
  interval_add_secondary(
    intervals_urine_only,
    param = "clr.last",
    reference = data.frame(PCSPEC = "plasma")
  )
)
```

The created plasma row requests `auclast`, which is what the link reads from it.

Because `reference` can name `start` and `end`, the same call links intervals in
time rather than across groups.  An accumulation ratio compares a dosing
interval with the first one:

```{r accumulation-ratio}
d_conc_acc <-
  data.frame(
    subject = 1,
    time = c(0, 6, 12, 24, 30, 36, 48),
    conc = c(0, 8, 5, 2, 12, 8, 3)
  )
d_dose_acc <- data.frame(subject = 1, time = c(0, 24), dose = 100)
intervals_acc <-
  interval_add_secondary(
    data.frame(start = c(0, 24), end = c(24, 48), aucint.last = TRUE),
    param = "ratio.aucint.last",
    reference = data.frame(start = 0, end = 24)
  )
kable_intervals(intervals_acc)
```

```{r accumulation-ratio-calculate}
o_data_acc <-
  PKNCAdata(
    PKNCAconc(d_conc_acc, conc~time|subject),
    PKNCAdose(d_dose_acc, dose~time|subject),
    intervals = intervals_acc,
    options = list(auc.method = "linear")
  )
kable(as.data.frame(pk.nca(o_data_acc)))
```

An interval that already names a different reference is left alone with a
warning, so the helper never silently re-points an analysis.

# Bioavailability

Bioavailability compares two administrations, so its reference is the other
treatment rather than another specimen.  Both the dose and the AUC come from the
reference interval:

```{r bioavailability}
d_conc_f <-
  data.frame(
    treatment = rep(c("ref", "test"), each = 5),
    subject = 1,
    time = rep(c(0, 1, 2, 4, 8), 2),
    conc = c(0, 10, 8, 5, 2, 0, 4, 3.5, 2, 0.8)
  )
d_dose_f <-
  data.frame(treatment = c("ref", "test"), subject = 1, time = 0, dose = c(100, 50))
intervals_f <-
  data.frame(
    treatment = c("ref", "test"),
    start = 0, end = Inf,
    interval_id = c("refprofile", NA),
    aucinf.obs = TRUE,
    totdose = TRUE,
    f.obs = c(FALSE, TRUE),
    f.obs_ref = c(NA, "refprofile")
  )
o_data_f <-
  PKNCAdata(
    PKNCAconc(d_conc_f, conc~time|treatment+subject),
    PKNCAdose(d_dose_f, dose~time|treatment+subject),
    intervals = intervals_f
  )
d_f <- as.data.frame(pk.nca(o_data_f))
kable(d_f[d_f$PPTESTCD %in% c("totdose", "aucinf.obs", "f.obs"), ])
```

The result is the ratio of the dose-normalized exposures:

```{r bioavailability-check}
value_of <- function(param, trt) {
  d_f$PPORRES[d_f$PPTESTCD %in% param & d_f$treatment %in% trt]
}
(value_of("aucinf.obs", "test")/value_of("totdose", "test")) /
  (value_of("aucinf.obs", "ref")/value_of("totdose", "ref"))
```

Bioavailability names the AUC it is built on, because which AUC was used changes
the answer and is not recoverable from the number:

```{r bioavailability-bases}
f_params <-
  c("f.obs", "f.pred", "f.last", "f.int.obs", "f.int.pred",
    "f.int.last", "f.int.all")
kable(
  data.frame(
    parameter = f_params,
    `AUC basis` =
      vapply(
        X = f_params,
        FUN = function(x) get.interval.cols()[[x]]$formalsmap$auc2,
        FUN.VALUE = ""
      ),
    check.names = FALSE
  ),
  row.names = FALSE
)
```

The `ratio.*` parameters are named the same way:  `ratio.cmax`,
`ratio.auclast`, `ratio.aucinf.obs`, `ratio.aucinf.pred`, `ratio.aucint.last`,
and `ratio.aucint.all` each compare the parameter they are named for.

# Exclusions cross the link

An exclusion on a value that feeds a secondary parameter reaches the result.
Here the plasma profile has too few points for a half-life, so the `aucinf.obs`
that extrapolates with it is excluded, and the renal clearance built on that AUC
inherits the reason:

```{r exclusion-data}
d_conc_x <-
  data.frame(
    subject = 1,
    PCSPEC = rep(c("plasma", "urine"), times = c(3, 2)),
    time = c(0, 2, 4, 0, 6),
    conc = c(10, 8, 7, 2, 1),
    vol  = c(NA, NA, NA, 100, 150)
  )
intervals_x <-
  data.frame(
    PCSPEC = c("plasma", "urine"),
    start = 0, end = 6,
    interval_id = c("plasma06", NA),
    aucinf.obs = c(TRUE, FALSE),
    ae = c(FALSE, TRUE),
    clr.obs = c(FALSE, TRUE),
    clr.obs_ref = c(NA, "plasma06")
  )
o_data_x <-
  PKNCAdata(
    PKNCAconc(d_conc_x, conc~time|PCSPEC+subject, volume = "vol"),
    intervals = intervals_x,
    options = list(auc.method = "linear")
  )
d_x <- as.data.frame(pk.nca(o_data_x))
kable(
  d_x[
    d_x$PPTESTCD %in% c("half.life", "aucinf.obs", "ae", "clr.obs"),
    c("PCSPEC", "PPTESTCD", "PPORRES", "exclude")
  ]
)
```

This is the property that post-hoc division of a results table does not have:
an excluded input cannot quietly become a reported clearance.

# Units that differ between the two profiles

A secondary result is reported in the units of the interval that requested it.
The reference value can be measured in different units -- a urine collection
reported in mg/L against a plasma profile in ng/mL -- and it is converted before
the calculation so that the reported units describe the result.

```{r units-data}
d_conc_u <- d_conc
d_conc_u$cu <- ifelse(d_conc_u$PCSPEC %in% "plasma", "ng/mL", "mg/L")
d_conc_u$tu <- "hr"
d_conc_u$au <- "mg"
o_conc_u <-
  PKNCAconc(
    d_conc_u, conc~time|PCSPEC+subject, volume = "vol",
    concu = "cu", timeu = "tu", amountu = "au"
  )
o_data_u <-
  PKNCAdata(
    o_conc_u, intervals = intervals_linked,
    options = list(auc.method = "linear")
  )
kable(as.data.frame(pk.nca(o_data_u)))
```

The plasma AUC is reported as 144 hr\*ng/mL on its own row.  Feeding the renal
clearance, it is 0.144 hr\*mg/L, and the clearance is
$350/0.144 = `r signif(350/(144*0.001), 6)`$ mg/(hr\*mg/L).  Conversion uses the
`units` package.

When the two units cannot be converted into one another -- outside udunits, or
of different dimensions -- the result is `NA` with the reason, and a
`pknca_warning_secondary_units` warning is raised once per parameter and unit
pair.  A number reported in units that do not describe it would be worse than no
number.

# Reporting

`PPANMETH` discloses the reference interval on every linked result:  how it
differs from the interval reporting the result, and its times.

```{r ppanmeth}
d_linked <- as.data.frame(results_linked)
d_linked$PPANMETH[d_linked$PPTESTCD %in% "clr.last"]
```

Secondary parameters are summarized like any other parameter, on the interval
that requested them.  Rows that did not request the parameter are `.`:

```{r summary}
kable(as.data.frame(summary(results_linked)))
```

`as.data.frame(filter_requested = TRUE)` reports only the parameters the
interval specification asked for, dropping the source values calculated to
support a link.

# Limitations

* Sparse concentration data (`PKNCAconc(sparse = TRUE)`) with any secondary
  parameter requested is an error.  Supporting it is planned.
* The reference for an absolute bioavailability is not yet derived from the
  route of administration; give it with `group_ref`, a `f.obs_ref` pointer, or
  `interval_add_secondary()`.
* The CDISC `PPTESTCD` of a ratio does not yet depend on what the ratio compares
  (accumulation against metabolite ratio).
* `exclude()` and the `exclude_nca_*()` functions work on a finished
  `PKNCAresults` and cannot reach across groups, so an exclusion applied after
  the calculation does not propagate along a link the way one set during the
  calculation does.
* Molecular-weight adjustment of a metabolite ratio (a molar conversion) is a
  units-layer feature that is not yet available.

# See also

* `vignette("v03-selection-of-calculation-intervals")` for interval
  specifications generally.
* `vignette("v07-unit-conversion")` for how units are assigned and converted.
* `vignette("v80-writing-parameter-functions")` for writing a parameter that
  needs another interval.
