---
title: "Sparse AUC to Infinity and the Uncertainty of Lambda z"
author: "Bill Denney"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Sparse AUC to Infinity and the Uncertainty of Lambda z}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(PKNCA)
simulation <- readRDS("v24-sparse-auc-to-infinity-simulation.rds")
# One summary value of the simulation, for the text
sim_value <- function(design, blq, elimination, method, parameter, column, digits = 1) {
  s <- simulation$summary
  row <-
    s$design == design & s$blq == blq & s$elimination == elimination &
    s$method == method & s$parameter == parameter
  stopifnot(sum(row) == 1)
  round(s[[column]][row], digits)
}
# The range of a summary value over scenarios
sim_range <- function(method, parameter, column, elimination = c("fast", "slow"), digits = 0) {
  s <- simulation$summary
  v <- s[[column]][s$method == method & s$parameter == parameter & s$elimination %in% elimination]
  paste(round(range(v), digits), collapse = " to ")
}
```

# Introduction

With sparse sampling, each animal contributes only one or a few samples to a
concentration-time profile, so there is no individual AUC to summarize.
Instead, the AUC is estimated from the mean concentration at each time, and its
standard error comes from the variability between animals at each time.  Bailer
(1988) gave the estimator and its variance for the AUC from time 0 to the last
sampling time with serial sacrifice (one sample per animal).  Nedelman and Jia
(1998) extended it to designs where animals give more than one sample, with
a Satterthwaite approximation for the degrees of freedom, and Holder (2001)
gave an unbiased estimator of the covariance between times.  Yuan (1993)
extended Bailer's method to the AUC from 0 to infinity by adding the
extrapolation $\bar{y}_n/\lambda_z$, treating the terminal elimination rate
$\lambda_z$ as known.

In practice, $\lambda_z$ is estimated from the same data, so its uncertainty
adds to that of the AUC to infinity.  This vignette describes how PKNCA
calculates the sparse AUC and AUMC to infinity with their standard errors and
degrees of freedom, gives a delta-method extension of Yuan's method that
includes the uncertainty in $\lambda_z$, and compares the methods in a
simulation-estimation study.

# The sparse AUC to the last measurement

At each sampling time $t_i$ ($i = 0, \ldots, n$), $r_i$ animals are sampled and
$\bar{y}_i$ is the arithmetic mean concentration.  $r_{ij}$ animals are sampled
at both $t_i$ and $t_j$ ($r_{ii} = r_i$).  The linear trapezoidal AUC of the
mean profile is a weighted sum of the means,

$$\widehat{AUC}_{0-t_{last}} = \sum_{i=0}^{n} w_i \bar{y}_i,
\quad w_0 = \frac{t_1 - t_0}{2},
\quad w_i = \frac{t_{i+1} - t_{i-1}}{2},
\quad w_n = \frac{t_n - t_{n-1}}{2},$$

where $t_n = t_{last}$ is the last time with a positive mean.  Since the
estimate is linear in the means, its variance is

$$\text{Var}\left(\widehat{AUC}_{0-t_{last}}\right) = \sum_{i}\sum_{j} w_i w_j \frac{r_{ij}\,\sigma_{ij}}{r_i r_j},$$

with $\sigma_{ij}$ the covariance of the concentrations at $t_i$ and $t_j$
within an animal.  PKNCA estimates $\sigma_{ii}$ with the sample variance and
$\sigma_{ij}$ ($i \neq j$) with the unbiased estimator of Holder (2001), setting
it to 0 when fewer than two animals are sampled at both times.  With serial
sacrifice, $r_{ij} = 0$ for $i \neq j$ and this is Bailer's (1988) variance,
$\sum_i w_i^2 s_i^2/r_i$.

Only times up to $t_{last}$ have weight.  When more than half of the samples at
a time are below the limit of quantification (BLQ), PKNCA sets that time's mean
to zero; times after the last positive mean are not part of
$AUC_{0-t_{last}}$, and they contribute nothing to its variance either.

# Extrapolation to infinity (Yuan 1993)

With first-order terminal elimination, the area after $t_{last}$ is
$\bar{y}_{n}/\lambda_z$, so

$$\widehat{AUC}_{0-\infty} = \sum_{i=0}^{n} w_i \bar{y}_i + \frac{\bar{y}_n}{\lambda_z}.$$

This is still linear in the means if $\lambda_z$ is known, with the last weight
increased by $1/\lambda_z$.  Yuan's (1993) variance (his equation 4) is
Bailer's with that weight, $g_n = w_n + 1/\lambda_z$:

$$\text{Var}\left(\widehat{AUC}_{0-\infty}\right) = \sum_{i}\sum_{j} g_i g_j \frac{r_{ij}\,\sigma_{ij}}{r_i r_j},
\qquad g_i = w_i + \frac{\delta_{i,n}}{\lambda_z}.$$

Written this way, the method applies to any sampling design, not only serial
sacrifice.

The AUMC is the AUC of the moment curve $t\,C(t)$.  Its estimate uses the
moment means $\bar{m}_i = t_i \bar{y}_i$, and the extrapolation
$t_{last} C_{last}/\lambda_z + C_{last}/\lambda_z^2$ is
$\bar{m}_n\left(1/\lambda_z + 1/(t_{last}\lambda_z^2)\right)$, so the AUMC
weights are $g_i = w_i + \delta_{i,n}\left(1/\lambda_z + 1/(t_{last}\lambda_z^2)\right)$
applied to the moment means, with the covariances of the moments
$t_i t_j \sigma_{ij}$.

# The uncertainty of lambda z (delta method)

$\lambda_z$ is estimated from the mean profile:  PKNCA fits a line to the log
mean concentrations at the half-life points, chosen automatically by the
adjusted $r^2$ (see `vignette("v06-half-life-calculation")`).  For a given
set of half-life points $H$, the estimate is minus the least-squares slope,

$$\hat{\lambda}_z = -\sum_{j \in H} c_j \log \bar{y}_j,
\qquad c_j = \frac{t_j - \bar{t}_H}{\sum_{k \in H}\left(t_k - \bar{t}_H\right)^2},$$

a smooth function of the means.  So $\widehat{AUC}_{0-\infty}$ is a smooth
function of the means, and the delta method gives its variance from the same
covariance of the means with the gradient in place of the weights:

$$g_j = \frac{\partial \widehat{AUC}_{0-\infty}}{\partial \bar{y}_j}
= w_j + \frac{\delta_{j,n}}{\lambda_z}
+ \frac{\bar{y}_n}{\lambda_z^2}\,\frac{c_j}{\bar{y}_j}\,\mathbb{1}_{j \in H}.$$

The last term is the uncertainty in $\lambda_z$.  Because the gradient is taken
with respect to all of the means together, the variance includes the covariance
of $\hat{\lambda}_z$ with $\bar{y}_n$ and with $\widehat{AUC}_{0-t_{last}}$.
Those covariances matter:  a high mean at an early half-life point raises the
AUC to $t_{last}$ but steepens the slope, which lowers the extrapolated area,
so the two partly cancel.  Adding a separately estimated variance of
$\hat{\lambda}_z$ to Yuan's variance would miss that.

For the AUMC, with the moment means,
$\hat{\lambda}_z = -\sum_{j \in H} c_j \left(\log \bar{m}_j - \log t_j\right)$ and

$$g_j = w_j + \delta_{j,n}\left(\frac{1}{\lambda_z} + \frac{1}{t_{last}\lambda_z^2}\right)
+ \bar{m}_n\left(\frac{1}{\lambda_z^2} + \frac{2}{t_{last}\lambda_z^3}\right)\frac{c_j}{\bar{m}_j}\,\mathbb{1}_{j \in H}.$$

The method conditions on the selected half-life points:  the variability from
choosing them is not included.  It is a first-order approximation, which is
most accurate when the coefficient of variation of the means at the half-life
points is small.

# Degrees of freedom

Confidence intervals use a $t$ distribution.  The variance estimate is a
quadratic form in the individual measurements $y$, $\hat{V} = y^T M y$, where
$M$ centers the measurements at each time and weights the products of each
animal's measurements by $g_i g_j r_{ij}/(r_i r_j h_{ij})$, with $h_{ij}$ the
divisor of the Holder covariance.  With $\Omega$ the covariance of the
measurements, Nedelman and Jia's (1998) equation 6 gives the Satterthwaite
degrees of freedom,

$$\nu = \frac{2\,E^2}{V}, \qquad E = \text{tr}(M\Omega), \qquad V = 2\,\text{tr}(M\Omega M\Omega).$$

With serial sacrifice, this is equation 6a of Nedelman, Gibiansky, and Lau
(1995).  PKNCA uses the same expression with the weights $g$ for the AUC and
AUMC to infinity, with or without the delta-method terms.

# An alternative:  lambda z from the individual samples

Instead of the log means, $\lambda_z$ could be estimated from every individual
sample in the terminal phase, with its standard error from that regression.
Samples below the limit of quantification are censored, so the regression is a
Tobit regression on the log concentrations; when animals give more than one
sample, a random intercept for each animal accounts for their correlation.  The
variance of $\hat{\lambda}_z$ is then added to Yuan's variance.

Two properties of this approach differ from the delta method.  A line through
the individual log concentrations follows the geometric-mean profile, while the
AUC and $C_{last}$ come from the arithmetic means; the two have the same
terminal slope only when the between-animal coefficient of variation is the
same at every terminal time.  And the regression does not give the covariance of
$\hat{\lambda}_z$ with the means, so it is added as though it were independent.
The simulation below compares both.

# Simulation-estimation study

```{r simulation-details, echo = FALSE}
model <- simulation$model
```

The study simulated a one-compartment model with first-order absorption (dose
`r model$dose`, typical $k_a$ `r model$ka`/hr and $V$ `r model$v`), log-normal
between-animal variability (`r 100*model$omega_ke`% on each of $k_a$, $k_e$,
and $V$), and a multiplicative residual error (`r 100*model$sigma`%) with mean
1.  Two elimination rates gave a small and a large extrapolated area, and two
limits of quantification gave no BLQ samples or about a third of the samples at
24 hours BLQ.  The serial-sacrifice design sampled `r simulation$n_serial`
animals at each of `r length(simulation$sample_times)` times
(`r paste(simulation$sample_times, collapse = ", ")` hours); the batch design
sampled three batches of four animals, each batch at three of those times.
Each of the `r nrow(simulation$scenarios)` scenarios had
`r format(simulation$n_replicates, big.mark = ",")` replicates.

The truth is the AUC (and AUMC) to infinity of the population mean
concentration curve, which is what the arithmetic means estimate; for this model
it is the mean of the individual $\text{dose}/(V k_e)$ (and
$\text{dose}/(V k_e)\,(1/k_e + 1/k_a)$), which has a closed form for log-normal
parameters.  For each replicate, the half-life was fit to the mean profile as
PKNCA does, and the three standard errors were calculated:

* **Yuan:**  $\lambda_z$ treated as known (`sparse_lambda_z_se = "none"`);
* **Delta:**  the delta method for $\lambda_z$ (`sparse_lambda_z_se = "delta"`);
* **Individual:**  $\lambda_z$ and its standard error from a Tobit regression
  of the individual samples from the first half-life time onward (with a random
  intercept for each animal in the batch design), with that $\lambda_z$ in the
  extrapolation and its variance added to Yuan's.

The simulation is in `data-raw/sparse_aucinf_simulation.R` in the PKNCA source,
which was run with PKNCA `r simulation$pknca_version` (`r simulation$r_version`,
`r simulation$run_date`); every replicate has its own seed, so the results can
be reproduced.

```{r simulation-results, echo = FALSE}
results <- simulation$summary
results$scenario_label <- paste(results$design, results$blq, results$elimination, sep = ", ")
results$method <- factor(results$method, levels = c("yuan", "delta", "individual"))
show_results <- function(parameter) {
  r <- results[results$parameter == parameter, ]
  r <- r[order(r$scenario, r$method), ]
  out <-
    data.frame(
      Scenario = r$scenario_label,
      Method = as.character(r$method),
      `Extrapolated (%)` = round(r$percent_extrapolated, 1),
      `Bias (%)` = round(r$relative_bias, 1),
      `RMSE (%)` = round(r$rmse, 1),
      `Median abs. error (%)` = round(r$median_absolute_error, 1),
      `Median 95% CI width (%)` = round(r$median_ci_width, 1),
      `Median df` = round(r$median_df, 1),
      `Coverage (%)` = round(r$coverage, 1),
      check.names = FALSE
    )
  knitr::kable(out, row.names = FALSE)
}
```

## AUC to infinity

```{r simulation-auc, echo = FALSE}
show_results("auc")
```

## AUMC to infinity

```{r simulation-aumc, echo = FALSE}
show_results("aumc")
```

Bias, root mean squared error (RMSE), median absolute error, and the median
width of the 95% confidence interval are relative to the true value.  Yuan and
delta share the point estimate, so they differ only in the standard error.  The
Monte Carlo standard error of each coverage is about
`r round(max(simulation$summary$coverage_mcse), 1)` percentage points or less.

## Findings

**Coverage of the AUC to infinity.**  When elimination was fast and 5 to 8% of
the AUC was extrapolated, all three methods covered close to the nominal 95%
(Yuan `r sim_range("yuan", "auc", "coverage", "fast")`%, delta
`r sim_range("delta", "auc", "coverage", "fast")`%, individual
`r sim_range("individual", "auc", "coverage", "fast")`%), so the uncertainty of
$\lambda_z$ matters little when little is extrapolated.  When elimination was
slow and about 22% was extrapolated, treating $\lambda_z$ as known (Yuan)
under-covered (`r sim_range("yuan", "auc", "coverage", "slow")`%), and the delta
method was closest to nominal in every scenario
(`r sim_range("delta", "auc", "coverage", "slow")`%).
Adding the variance of $\lambda_z$ from the individual samples
(`r sim_range("individual", "auc", "coverage", "slow")`% with slow elimination)
recovered only part of the shortfall, because it omits the covariance of
$\hat{\lambda}_z$ with the means.

**Precision of the point estimate.**  The median absolute errors were similar
for the two estimates of $\lambda_z$, but the log-mean $\lambda_z$ occasionally
came out near zero when samples were BLQ or elimination was slow, which made
the AUC to infinity heavy-tailed:  with serial sacrifice, BLQ samples, and fast
elimination, the RMSE was `r sim_value("serial", "moderate", "fast", "delta", "auc", "rmse")`%
with the log-mean $\lambda_z$ and
`r sim_value("serial", "moderate", "fast", "individual", "auc", "rmse")`% with
the Tobit $\lambda_z$ from the individual samples.  The Tobit fit uses every
terminal sample, including the censored ones, rather than means that count BLQ
samples as zero.  That advantage was not uniform:  in the batch design with BLQ
samples and slow elimination, the mixed-effects Tobit fit was itself unstable
(RMSE `r sim_value("batch", "moderate", "slow", "individual", "auc", "rmse")`%,
against `r sim_value("batch", "moderate", "slow", "delta", "auc", "rmse")`%).

**The AUMC to infinity.**  All three methods under-covered the AUMC
(delta `r sim_range("delta", "aumc", "coverage")`%, Yuan
`r sim_range("yuan", "aumc", "coverage")`%), more so with slow elimination.  Its
extrapolation is proportional to $1/\lambda_z^2$, so the first-order delta
method is a poorer approximation, and the estimate is more heavy-tailed.

**Interval width.**  The delta-method intervals were wider than Yuan's, by the
most when elimination was slow; the extra width is what improved the coverage.

# Calculating it with PKNCA

Request `aucinf.obs` (or `aumcinf.obs`) with sparse data.  The standard error
and degrees of freedom are reported with it as `aucinf.obs_se` and
`aucinf.obs_df`, and the summary shows the estimate with its standard error.

```{r pknca-example}
d_sparse <-
  data.frame(
    time = rep(c(0, 0.5, 1, 2, 4, 6, 8, 12, 24), each = 4),
    conc =
      c(0, 0, 0, 0,  3.1, 4.4, 3.6, 5.2,  5.8, 6.9, 5.1, 7.4,  7.2, 6.1, 8.3, 6.6,
        5.0, 6.2, 4.4, 5.6,  4.1, 3.4, 4.8, 3.7,  2.9, 3.5, 2.4, 3.1,
        1.6, 1.9, 1.3, 1.8,  0.33, 0.27, 0.41, 0.30)
  )
d_sparse$animal <- seq_len(nrow(d_sparse))
o_data <-
  PKNCAdata(
    PKNCAconc(d_sparse, conc ~ time | animal, sparse = TRUE),
    intervals = data.frame(start = 0, end = Inf, aucinf.obs = TRUE, aumcinf.obs = TRUE),
    options = list(sparse_lambda_z_se = "delta")
  )
o_nca <- pk.nca(o_data)
d_result <- as.data.frame(o_nca)
d_result[grepl("^au.*inf|^lambda.z$", d_result$PPTESTCD), c("PPTESTCD", "PPORRES")]
summary(o_nca)
```

The `sparse_lambda_z_se` option (set globally with `PKNCA.options()` or for
one analysis with the `options` argument of `PKNCAdata()`) is `"none"` for
Yuan's method or `"delta"` for the delta method.  The method used is recorded in
the `PPANMETH` column.

# Discussion

When $\lambda_z$ is estimated from the same samples, its uncertainty adds to
that of the AUC to infinity, and treating it as known, as Yuan's (1993)
standard error does, under-covers once the extrapolated area is a substantial
part of the AUC.  The delta method on the log means is a consistent extension of
Yuan's method:  it keeps his estimate, applies to serial-sacrifice and batch
designs alike through the covariance of the means, includes the covariance of
$\hat{\lambda}_z$ with $C_{last}$ and the AUC to $t_{last}$, and needs nothing
beyond the quantities already used for the sparse AUC to $t_{last}$.  In the
simulation it brought the coverage of the AUC to infinity much closer to
nominal when the extrapolated area was large
(`r sim_range("delta", "auc", "coverage", "slow")`%, against
`r sim_range("yuan", "auc", "coverage", "slow")`% for Yuan), though still a few
points short.  For this reason, PKNCA offers it as the extension of Yuan's method
(`sparse_lambda_z_se = "delta"`).

Estimating $\lambda_z$ from the individual samples with a (mixed-effects) Tobit
regression gave a more stable point estimate when samples were BLQ, but adding
its variance to Yuan's did not give adequate coverage, and its mixed-effects
fit was unstable in the sparsest design.  A natural refinement, not studied
here, would combine the individual-sample $\lambda_z$ with a variance that
includes its covariance with the means, for example by a joint delta method or a
bootstrap over animals.

Limitations remain.  The delta method conditions on the selected half-life
points, so the uncertainty of the automatic point selection is not included; it
is a first-order approximation, which is least accurate when $\lambda_z$ is
imprecise, as for the AUMC with slow elimination; and the arithmetic mean
profile of animals with different elimination rates is not monoexponential, so
the extrapolation $C_{last}/\lambda_z$ is itself biased for the population mean
curve.  Where the extrapolated area is large, sampling longer remains the
better remedy.

# References

Bailer AJ.  Testing for the equality of area under the curves when using
destructive measurement techniques.  Journal of Pharmacokinetics and
Biopharmaceutics.  1988;16(3):303-309.

Holder DJ.  Comments on Nedelman and Jia's extension of Satterthwaite's
approximation applied to pharmacokinetics.  Journal of Biopharmaceutical
Statistics.  2001;11(1-2):75-79.  doi:10.1081/BIP-100104199

Nedelman JR, Gibiansky E, Lau DTW.  Applying Bailer's method for AUC
confidence intervals to sparse sampling.  Pharmaceutical Research.
1995;12(1):124-128.

Nedelman JR, Jia X.  An extension of Satterthwaite's approximation applied to
pharmacokinetics.  Journal of Biopharmaceutical Statistics.
1998;8(2):317-328.  doi:10.1080/10543409808835241

Yuan J.  Estimation of variance for AUC in animal studies.  Journal of
Pharmaceutical Sciences.  1993;82(7):761-763.  doi:10.1002/jps.2600820718
