---
title: "Sparse AUC to Infinity and the Uncertainty of lambda.z"
author: "Bill Denney"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Sparse AUC to Infinity and the Uncertainty of lambda.z}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(PKNCA)
simulation <- readRDS("v24-sparse-auc-to-infinity-simulation.rds")
mean_type <- readRDS("v24-sparse-auc-to-infinity-mean-type.rds")
bootstrap <- readRDS("v24-sparse-auc-to-infinity-bootstrap.rds")
bootstrap_n <- readRDS("v24-sparse-auc-to-infinity-bootstrap-n.rds")
# The rows of a simulation summary that match every given column value
sim_rows <- function(summary, filters) {
  keep <- rep(TRUE, nrow(summary))
  for (current_name in names(filters)) {
    keep <- keep & summary[[current_name]] %in% filters[[current_name]]
  }
  summary[keep, ]
}
# The range of a column of the standard error study over the rows matching
# `...` (column = value), as text
se_range <- function(column, ..., digits = 0) {
  v <- sim_rows(simulation$summary, list(...))[[column]]
  paste(format(round(range(v), digits), nsmall = digits), collapse = " to ")
}
# The same for the bootstrap study
boot_range <- function(column, ..., digits = 0) {
  v <- sim_rows(bootstrap$summary, list(...))[[column]]
  paste(format(round(range(v), digits), nsmall = digits), collapse = " to ")
}
# One value of the profile over the number of animals per time
profile_value <- function(column, n_per_time, blq, method, parameter, digits = 1) {
  v <- sim_rows(bootstrap_n$summary, list(n_per_time = n_per_time, blq = blq, method = method, parameter = parameter))[[column]]
  stopifnot(length(v) == 1)
  format(round(v, digits), nsmall = digits)
}
# The same for the arithmetic and geometric mean profile study
mt_range <- function(column, ..., digits = 0) {
  v <- sim_rows(mean_type$summary, list(...))[[column]]
  paste(format(round(range(v), digits), nsmall = digits), collapse = " to ")
}
```

# Summary and recommendations

For sparse sampling, where each animal gives one or a few samples:

* **Use the delta-method standard error for the AUC to infinity (the
  default).**  When little of the AUC is extrapolated, Yuan's standard error,
  which treats $\lambda_z$ as known, and the delta method both cover close to
  the nominal 95% (Yuan `r se_range("coverage", method = "yuan", parameter = "auc", elimination = "fast")`%).
  When about a fifth of the AUC is extrapolated, Yuan's intervals are too
  narrow (`r se_range("coverage", method = "yuan", parameter = "auc", elimination = "slow")`%
  coverage), and the delta method is closer to nominal
  (`r se_range("coverage", method = "delta", parameter = "auc", elimination = "slow")`%).
  Set `sparse_lambda_z_se = "none"` for Yuan's method.
* **Give the AUC a concentration at time 0.**  As for any PKNCA AUC, the sparse
  AUC is `NA` without one.  Sacrificial designs rarely sample at the dose, so
  after an extravascular dose impute zero with `impute = "start_conc0"`.  After
  an IV bolus, request the IV parameters (`aucivlast`, `aucivinf.obs`):  they
  back-extrapolate $C_0$ from the mean profile and include its uncertainty.
* **Treat the AUMC, and the MRT and Vss that come from it, with caution.**  The
  AUMC to infinity was biased low in most scenarios, and its confidence intervals
  under-covered in every scenario, even with the delta method
  (`r se_range("coverage", method = "delta", parameter = "aumc")`%), most when
  elimination was slow.
* **The half-life of the mean profile is not the typical animal's half-life.**
  It was too long when the terminal phase was well sampled (by up to
  `r max(round(sim_rows(mean_type$summary, list(mean_type = "arithmetic", parameter = "half.life", elimination = "fast", blq = "none"))$bias_typical))`%)
  and too short when the profile was cut short by BLQ samples or a two-compartment
  distribution (down to
  `r min(round(sim_rows(mean_type$summary, list(mean_type = "arithmetic", parameter = "half.life"))$bias_typical))`%).
  Sample long enough to see at least three terminal half-lives.
* **Cmax of the mean profile reads high with few animals per time.**  Taking
  the maximum of noisy means near a broad peak overestimated Cmax by up to
  `r max(round(sim_rows(mean_type$summary, list(mean_type = "arithmetic", parameter = "cmax", route = "oral"))$bias_natural))`%.
* **Geometric-mean profiles were evaluated and not adopted.**  They were no more
  biased and were as precise or modestly more precise, but they estimate a
  different quantity (the typical animal rather than the population mean
  curve), need a rule for BLQ samples that drives their BLQ results, and fall
  outside the variance methods for sparse AUCs, which are for arithmetic means;
  adopting them would change all sparse calculations for a modest gain.
* **The bootstrap (`sparse_bootstrap()`) gives intervals for any parameter, but
  they are too narrow with few animals at each time.**  With three or four
  animals per time, its 95% intervals for the AUC to infinity covered
  `r boot_range("coverage_percentile", parameter = "auc")`% of the time.  Prefer
  the analytical standard errors for the AUCs, and use the bootstrap for
  parameters and comparisons that have none, such as Cmax and ratios between
  groups.  With 20 or more animals per time, its standard error was accurate
  and its intervals matched the delta method's.
* **With many animals, the bias of the sparse estimate dominates.**  BLQ
  samples and the extrapolation of the mean profile biased the AUC to infinity
  low (by about
  `r gsub("-", "", profile_value("median_bias", 100, "moderate", "yuan", "auc", digits = 0))`%
  with BLQ samples here), whatever the number of animals, so with many animals
  every confidence interval missed the truth more often than its nominal
  level.
* **Estimating $\lambda_z$ from the individual samples was evaluated and not
  adopted.**  A (mixed-effects) Tobit fit gave a steadier point estimate when
  $\lambda_z$ was uncertain, but its standard error, added to Yuan's, did not
  cover as well as the delta method.

# Introduction

With sparse sampling, each animal contributes only one or a few samples to a
concentration-time profile, so there is no individual AUC to summarize.
Instead, the AUC is estimated from the mean concentration at each time, and its
standard error comes from the variability between animals at each time.  Bailer
(1988) gave the estimator and its variance for the AUC from time 0 to the last
sampling time with serial sacrifice (one sample per animal).  Nedelman and Jia
(1998) extended it to designs where animals give more than one sample, with
a Satterthwaite approximation for the degrees of freedom, and Holder (2001)
gave an unbiased estimator of the covariance between times.  Yuan (1993)
extended Bailer's method to the AUC from 0 to infinity by adding the
extrapolation $\bar{y}_n/\lambda_z$, treating the terminal elimination rate
$\lambda_z$ as known.

In practice, $\lambda_z$ is estimated from the same data, so its uncertainty
adds to that of the AUC to infinity.  This vignette describes how PKNCA
calculates the sparse AUC and AUMC to infinity with their standard errors and
degrees of freedom, gives a delta-method extension of Yuan's method that
includes the uncertainty in $\lambda_z$, extends both to IV bolus dosing, and
compares the methods in a simulation-estimation study.  A second study compares
arithmetic and geometric mean profiles.

# The sparse AUC to the last measurement

At each sampling time $t_i$ ($i = 0, \ldots, n$), $r_i$ animals are sampled and
$\bar{y}_i$ is the arithmetic mean concentration.  $r_{ij}$ animals are sampled
at both $t_i$ and $t_j$ ($r_{ii} = r_i$).  The linear trapezoidal AUC of the
mean profile is a weighted sum of the means,

$$\widehat{AUC}_{0-t_{last}} = \sum_{i=0}^{n} w_i \bar{y}_i,
\quad w_0 = \frac{t_1 - t_0}{2},
\quad w_i = \frac{t_{i+1} - t_{i-1}}{2},
\quad w_n = \frac{t_n - t_{n-1}}{2},$$

where $t_n = t_{last}$ is the last time with a positive mean.  Since the
estimate is linear in the means, its variance is

$$\text{Var}\left(\widehat{AUC}_{0-t_{last}}\right) = \sum_{i}\sum_{j} w_i w_j \frac{r_{ij}\,\sigma_{ij}}{r_i r_j},$$

with $\sigma_{ij}$ the covariance of the concentrations at $t_i$ and $t_j$
within an animal.  PKNCA estimates $\sigma_{ii}$ with the sample variance and
$\sigma_{ij}$ ($i \neq j$) with the unbiased estimator of Holder (2001), setting
it to 0 when fewer than two animals are sampled at both times.  With serial
sacrifice, $r_{ij} = 0$ for $i \neq j$ and this is Bailer's (1988) variance,
$\sum_i w_i^2 s_i^2/r_i$.

Only times up to $t_{last}$ have weight.  When more than half of the samples at
a time are below the limit of quantification (BLQ), PKNCA sets that time's mean
to zero; times after the last positive mean are not part of
$AUC_{0-t_{last}}$, and they contribute nothing to its variance either.

As for any AUC in PKNCA, the area starts at the start of the interval, so a
concentration is needed there, measured or imputed; without one, the sparse
AUC is `NA`.  Animals are rarely sampled at the dose in sacrificial designs, so
after an extravascular dose the usual choice is to impute a concentration of
zero with `impute = "start_conc0"`.  An imputed zero is known rather than
measured, so it has weight in the estimate (as $\bar{y}_0 = 0$) but no
variance.

# Extrapolation to infinity (Yuan 1993)

With first-order terminal elimination, the area after $t_{last}$ is
$\bar{y}_{n}/\lambda_z$, so

$$\widehat{AUC}_{0-\infty} = \sum_{i=0}^{n} w_i \bar{y}_i + \frac{\bar{y}_n}{\lambda_z}.$$

This is still linear in the means if $\lambda_z$ is known, with the last weight
increased by $1/\lambda_z$.  Yuan's (1993) variance (his equation 4) is
Bailer's with that weight, $g_n = w_n + 1/\lambda_z$:

$$\text{Var}\left(\widehat{AUC}_{0-\infty}\right) = \sum_{i}\sum_{j} g_i g_j \frac{r_{ij}\,\sigma_{ij}}{r_i r_j},
\qquad g_i = w_i + \frac{\delta_{i,n}}{\lambda_z}.$$

Written this way, the method applies to any sampling design, not only serial
sacrifice.

The AUMC is the AUC of the moment curve $t\,C(t)$.  Its estimate uses the
moment means $\bar{m}_i = t_i \bar{y}_i$, and the extrapolation
$t_{last} C_{last}/\lambda_z + C_{last}/\lambda_z^2$ is
$\bar{m}_n\left(1/\lambda_z + 1/(t_{last}\lambda_z^2)\right)$, so the AUMC
weights are $g_i = w_i + \delta_{i,n}\left(1/\lambda_z + 1/(t_{last}\lambda_z^2)\right)$
applied to the moment means, with the covariances of the moments
$t_i t_j \sigma_{ij}$.

# IV bolus:  the back-extrapolated C0

After an IV bolus, the concentration at time 0 is not zero, and it cannot be
measured.  As for dense data, PKNCA estimates it from the mean profile with the
methods of `pk.calc.c0()`:  log-linear back-extrapolation from the first two
means when they decline,

$$\hat{C}_0 = \bar{y}_1 \left(\frac{\bar{y}_1}{\bar{y}_2}\right)^{t_1/(t_2 - t_1)},$$

and otherwise $\bar{y}_1$.  The AUC adds the trapezoid from time 0 to $t_1$,
$t_1(\hat{C}_0 + \bar{y}_1)/2$.  $\hat{C}_0$ is a smooth function of the
first two means, so its uncertainty enters the variance through its gradient,
as in the next section:  the weights of $\bar{y}_1$ and $\bar{y}_2$ gain
$(t_1/2)\,\partial \hat{C}_0/\partial \bar{y}_j$, with

$$\frac{\partial \hat{C}_0}{\partial \bar{y}_1} = \frac{\hat{C}_0\, t_2}{(t_2 - t_1)\,\bar{y}_1},
\qquad
\frac{\partial \hat{C}_0}{\partial \bar{y}_2} = -\frac{\hat{C}_0\, t_1}{(t_2 - t_1)\,\bar{y}_2}.$$

With the linear trapezoidal rule, the AUMC from time 0 to $t_1$ is
$t_1^2\bar{y}_1/2$ whatever $C_0$ is, so $C_0$ does not affect the AUMC.  These
are the sparse `aucivlast`, `aucivall`, and `aucivinf.obs` (and their AUMC
equivalents), each with its standard error and degrees of freedom.

# The uncertainty of lambda.z (delta method)

$\lambda_z$ is estimated from the mean profile:  PKNCA fits a line to the log
mean concentrations at the half-life points, chosen automatically by the
adjusted $r^2$ (see `vignette("v06-half-life-calculation")`).  For a given
set of half-life points $H$, the estimate is minus the least-squares slope,

$$\hat{\lambda}_z = -\sum_{j \in H} c_j \log \bar{y}_j,
\qquad c_j = \frac{t_j - \bar{t}_H}{\sum_{k \in H}\left(t_k - \bar{t}_H\right)^2},$$

a smooth function of the means.  So $\widehat{AUC}_{0-\infty}$ is a smooth
function of the means, and the delta method gives its variance from the same
covariance of the means with the gradient in place of the weights:

$$g_j = \frac{\partial \widehat{AUC}_{0-\infty}}{\partial \bar{y}_j}
= w_j + \frac{\delta_{j,n}}{\lambda_z}
+ \frac{\bar{y}_n}{\lambda_z^2}\,\frac{c_j}{\bar{y}_j}\,\mathbb{1}_{j \in H}.$$

The last term is the uncertainty in $\lambda_z$.  Because the gradient is taken
with respect to all of the means together, the variance includes the covariance
of $\hat{\lambda}_z$ with $\bar{y}_n$ and with $\widehat{AUC}_{0-t_{last}}$.
Those covariances matter:  a high mean at an early half-life point raises the
AUC to $t_{last}$ but steepens the slope, which lowers the extrapolated area,
so the two partly cancel.  Adding a separately estimated variance of
$\hat{\lambda}_z$ to Yuan's variance would miss that.

For the AUMC, with the moment means,
$\hat{\lambda}_z = -\sum_{j \in H} c_j \left(\log \bar{m}_j - \log t_j\right)$ and

$$g_j = w_j + \delta_{j,n}\left(\frac{1}{\lambda_z} + \frac{1}{t_{last}\lambda_z^2}\right)
+ \bar{m}_n\left(\frac{1}{\lambda_z^2} + \frac{2}{t_{last}\lambda_z^3}\right)\frac{c_j}{\bar{m}_j}\,\mathbb{1}_{j \in H}.$$

The method conditions on the selected half-life points:  the variability from
choosing them is not included.  It is a first-order approximation, which is
most accurate when the coefficient of variation of the means at the half-life
points is small.

# Degrees of freedom

Confidence intervals use a $t$ distribution.  The variance estimate is a
quadratic form in the individual measurements $y$, $\hat{V} = y^T M y$, where
$M$ centers the measurements at each time and weights the products of each
animal's measurements by $g_i g_j r_{ij}/(r_i r_j h_{ij})$, with $h_{ij}$ the
divisor of the Holder covariance.  With $\Omega$ the covariance of the
measurements, Nedelman and Jia's (1998) equation 6 gives the Satterthwaite
degrees of freedom,

$$\nu = \frac{2\,E^2}{V}, \qquad E = \text{tr}(M\Omega), \qquad V = 2\,\text{tr}(M\Omega M\Omega).$$

With serial sacrifice, this is equation 6a of Nedelman, Gibiansky, and Lau
(1995).  PKNCA uses the same expression with the weights $g$ for the AUC and
AUMC to infinity, with or without the delta-method terms.

# An alternative:  lambda.z from the individual samples

Instead of the log means, $\lambda_z$ could be estimated from every individual
sample in the terminal phase, with its standard error from that regression.
Samples below the limit of quantification are censored, so the regression is a
Tobit regression on the log concentrations; when animals give more than one
sample, a random intercept for each animal accounts for their correlation.  The
variance of $\hat{\lambda}_z$ is then added to Yuan's variance.

Two properties of this approach differ from the delta method.  A line through
the individual log concentrations follows the geometric-mean profile, while the
AUC and $C_{last}$ come from the arithmetic means; the two have the same
terminal slope only when the between-animal coefficient of variation is the
same at every terminal time.  And the regression does not give the covariance of
$\hat{\lambda}_z$ with the means, so it is added as though it were independent.
The simulation below compares both.

# Simulation-estimation study

```{r simulation-details, echo = FALSE}
model <- simulation$model
```

## Design

Both studies in this vignette share one design
(`data-raw/sparse_simulation_design.R` in the PKNCA source).  Concentrations come
from one- and two-compartment models (pmxTools) after a dose of
`r model$dose`, given orally or as an IV bolus, with log-normal between-animal
variability (`r 100*model$omega`% on each structural parameter) and a
multiplicative residual error (`r 100*model$sigma`%) with mean 1:

* **One compartment:**  $V$ `r simulation$model_1cmt$v`, $k_e$ 0.15/hr (fast
  elimination, half-life 4.6 hr) or 0.07/hr (slow, 9.9 hr), and oral $k_a$
  `r simulation$model_1cmt$ka`/hr (Tmax 0.7 to 0.9 hr).
* **Two compartments:**  $V_1$ `r simulation$model_2cmt$v1`, $V_2$
  `r simulation$model_2cmt$v2`, $Q$ `r simulation$model_2cmt$q`, and $CL$
  chosen for the same terminal half-lives after a distribution phase with a
  half-life of about 0.5 hr; oral $k_a$ `r simulation$model_2cmt$ka`/hr (Tmax
  0.6 to 0.7 hr).

The slow elimination leaves about a fifth of the AUC to extrapolate after 24
hours, and the fast elimination less than a tenth.  Two limits of
quantification gave no BLQ samples or about a third of the samples at 24 hours
BLQ (70% of the typical concentration at 24 hours).  No animal was sampled at
the dose.  The serial-sacrifice design sampled `r simulation$n_serial` animals
at each of `r length(simulation$sampling_times$oral)` times
(`r paste(signif(simulation$sampling_times$oral, 2), collapse = ", ")` hours
for oral dosing; for IV bolus dosing, the first time was 5 minutes and 6 hours
was omitted); the batch design sampled three batches of
`r simulation$n_batch` animals, each batch at a third of those times.  With
`r nrow(simulation$scenarios)` scenarios (two models, two routes, two designs,
two limits of quantification, and two elimination rates), the standard error
study had `r format(simulation$n_replicates, big.mark = ",")` replicates per
scenario.

The truth is the AUC (and AUMC) from 0 to infinity of the population mean
concentration curve, which is what the arithmetic means estimate:  the mean of
the individual values over `r format(simulation$n_virtual, big.mark = ",")`
virtual animals (Monte Carlo standard error at most
`r round(max(simulation$truth_mcse_percent[, c("auc", "aumc")]), 2)`%).  For
each replicate, the oral concentration at time 0 was imputed as zero and the IV
bolus $C_0$ was back-extrapolated, the half-life was fit to the mean profile as
PKNCA does, and three standard errors were calculated:

* **Yuan:**  $\lambda_z$ treated as known (`sparse_lambda_z_se = "none"`);
* **Delta:**  the delta method for $\lambda_z$ (`sparse_lambda_z_se = "delta"`);
* **Individual:**  $\lambda_z$ and its standard error from a Tobit regression
  of the individual samples from the first half-life time onward (with a random
  intercept for each animal in the batch design), with that $\lambda_z$ in the
  extrapolation and its variance added to Yuan's.

The simulation is in `data-raw/sparse_aucinf_simulation.R` in the PKNCA source,
which was run with PKNCA `r simulation$pknca_version` and pmxTools
`r simulation$pmxtools_version` (`r simulation$r_version`,
`r simulation$run_date`); every replicate has its own seed, so the results can
be reproduced.

```{r simulation-results, echo = FALSE}
results <- simulation$summary
scenario_columns <- function(r) {
  data.frame(
    Model = paste0(r$compartments, "-cmt"),
    Route = ifelse(r$route == "iv", "IV bolus", "oral"),
    Design = r$design,
    BLQ = r$blq,
    Elimination = r$elimination
  )
}
# One row per scenario with a column of `column` for each method
by_method <- function(parameter, column, methods, digits = 1) {
  r <- results[results$parameter == parameter, ]
  scenarios <- sort(unique(r$scenario))
  ret <- data.frame(scenario = scenarios)
  for (current_method in methods) {
    current <- r[r$method == current_method, ]
    ret[[current_method]] <- round(current[[column]][match(scenarios, current$scenario)], digits)
  }
  ret
}
show_coverage <- function(parameter) {
  r <- results[results$parameter == parameter & results$method == "yuan", ]
  r <- r[order(r$scenario), ]
  coverage <- by_method(parameter, "coverage", c("yuan", "delta", "individual"))
  width <- by_method(parameter, "median_ci_width", c("yuan", "delta", "individual"), digits = 0)
  out <-
    data.frame(
      scenario_columns(r),
      `Extrap. (%)` = round(r$percent_extrapolated, 1),
      `Coverage Yuan` = coverage$yuan,
      `Coverage delta` = coverage$delta,
      `Coverage indiv.` = coverage$individual,
      `CI width Yuan` = width$yuan,
      `CI width delta` = width$delta,
      `CI width indiv.` = width$individual,
      check.names = FALSE
    )
  knitr::kable(out, row.names = FALSE)
}
show_accuracy <- function(parameter) {
  r <- results[results$parameter == parameter & results$method == "yuan", ]
  r <- r[order(r$scenario), ]
  bias <- by_method(parameter, "relative_bias", c("yuan", "individual"))
  mae <- by_method(parameter, "median_absolute_error", c("yuan", "individual"))
  rmse <- by_method(parameter, "rmse", c("yuan", "individual"))
  out <-
    data.frame(
      scenario_columns(r),
      `Bias mean` = bias$yuan,
      `Bias indiv.` = bias$individual,
      `MAE mean` = mae$yuan,
      `MAE indiv.` = mae$individual,
      `RMSE mean` = rmse$yuan,
      `RMSE indiv.` = rmse$individual,
      check.names = FALSE
    )
  knitr::kable(out, row.names = FALSE)
}
```

## Results:  coverage of the AUC to infinity

```{r simulation-auc-coverage, echo = FALSE}
show_coverage("auc")
```

## Results:  coverage of the AUMC to infinity

```{r simulation-aumc-coverage, echo = FALSE}
show_coverage("aumc")
```

Coverage is the percentage of 95% confidence intervals that contain the truth,
and the CI width is the median width relative to the truth (%).  The Monte Carlo
standard error of each coverage is at most
`r round(max(simulation$summary$coverage_mcse), 1)` percentage points.

## Results:  the bootstrap

The bootstrap of Shen and Machado (2017) (`sparse_bootstrap()`) was run on the
first `r format(bootstrap$n_replicates, big.mark = ",")` data sets of each
scenario with `r bootstrap$n_boot` replicates each, its default.  Each replicate
resampled the animals within each sampling time (or batch), and its AUC and
AUMC to infinity were calculated as for the original data.  Two intervals were
formed:  the percentile interval of the replicates and the estimate plus or
minus 1.96 bootstrap standard errors (the standard deviation of the
replicates).  The delta method (on all of the scenario's data sets, which
include these) is shown for comparison.

```{r simulation-bootstrap, echo = FALSE}
show_bootstrap <- function(parameter) {
  b <- bootstrap$summary[bootstrap$summary$parameter == parameter, ]
  b <- b[order(b$scenario), ]
  delta <- results[results$parameter == parameter & results$method == "delta", ]
  delta <- delta[match(b$scenario, delta$scenario), ]
  out <-
    data.frame(
      scenario_columns(b),
      `Coverage percentile` = round(b$coverage_percentile, 1),
      `Coverage boot. SE` = round(b$coverage_normal, 1),
      `Coverage delta` = round(delta$coverage, 1),
      `CI width percentile` = round(b$median_ci_width_percentile),
      `CI width delta` = round(delta$median_ci_width),
      check.names = FALSE
    )
  knitr::kable(out, row.names = FALSE)
}
```

```{r simulation-bootstrap-auc, echo = FALSE}
show_bootstrap("auc")
```

```{r simulation-bootstrap-aumc, echo = FALSE}
show_bootstrap("aumc")
```

## Results:  the bootstrap with more animals at each time

Published uses of the bootstrap had either a few animals per time (three mice in
Takemoto et al. 2006, which compared only the means and standard deviations
with Bailer's) or many (about 88 subjects per time and treatment in the
bioequivalence study of Shen and Machado 2017).  To see how the number of
animals per time changes the comparison, the two-compartment, oral,
serial-sacrifice design with slow elimination was simulated with
`r paste(bootstrap_n$n_per_time, collapse = ", ")` animals per time, without
and with BLQ samples (`r format(bootstrap_n$n_replicates, big.mark = ",")` data
sets each, `r bootstrap_n$n_boot` bootstrap replicates;
`data-raw/sparse_bootstrap_n_profile.R`).  The data seeds do not depend on the
number of animals, so 4 animals per time reproduces the data sets above.  The
bias is that of the point estimate, which all of the methods share.

```{r simulation-bootstrap-n, echo = FALSE}
show_bootstrap_n <- function(parameter) {
  b <- bootstrap_n$summary[bootstrap_n$summary$parameter == parameter, ]
  keys <- unique(b[, c("blq", "n_per_time")])
  keys <- keys[order(keys$blq != "none", keys$n_per_time), ]
  out <- data.frame()
  for (i in seq_len(nrow(keys))) {
    current <- merge(b, keys[i, ])
    value_of <- current[match(c("percentile", "boot_se", "yuan", "delta"), current$method), ]
    out <-
      rbind(
        out,
        data.frame(
          BLQ = keys$blq[i],
          `Animals per time` = keys$n_per_time[i],
          `Coverage percentile` = round(value_of$coverage[1], 1),
          `Coverage boot. SE` = round(value_of$coverage[2], 1),
          `Coverage Yuan` = round(value_of$coverage[3], 1),
          `Coverage delta` = round(value_of$coverage[4], 1),
          `CI width percentile` = round(value_of$median_ci_width[1]),
          `CI width delta` = round(value_of$median_ci_width[4]),
          `Bias (%)` = round(value_of$median_bias[1], 1),
          check.names = FALSE
        )
      )
  }
  knitr::kable(out, row.names = FALSE)
}
```

```{r simulation-bootstrap-n-auc, echo = FALSE}
show_bootstrap_n("auc")
```

```{r simulation-bootstrap-n-aumc, echo = FALSE}
show_bootstrap_n("aumc")
```

## Results:  accuracy of the point estimates

The point estimate with $\lambda_z$ from the log means ("mean", shared by Yuan
and delta) and with $\lambda_z$ from the Tobit fit of the individual samples
("indiv."):  bias, median absolute error (MAE), and root mean squared error
(RMSE), all relative to the truth (%).

```{r simulation-auc-accuracy, echo = FALSE}
show_accuracy("auc")
```

```{r simulation-aumc-accuracy, echo = FALSE}
show_accuracy("aumc")
```

## Findings

**Coverage of the AUC to infinity.**  With fast elimination, when
`r se_range("percent_extrapolated", parameter = "auc", method = "yuan", elimination = "fast")`%
of the AUC was extrapolated, all three methods covered close to the nominal 95%
(Yuan `r se_range("coverage", method = "yuan", parameter = "auc", elimination = "fast", digits = 1)`%,
delta `r se_range("coverage", method = "delta", parameter = "auc", elimination = "fast", digits = 1)`%,
individual `r se_range("coverage", method = "individual", parameter = "auc", elimination = "fast", digits = 1)`%),
so the uncertainty of $\lambda_z$ matters little when little is extrapolated.
With slow elimination, when
`r se_range("percent_extrapolated", parameter = "auc", method = "yuan", elimination = "slow")`%
was extrapolated, treating $\lambda_z$ as known under-covered (Yuan
`r se_range("coverage", method = "yuan", parameter = "auc", elimination = "slow", digits = 1)`%),
and the delta method was closest to nominal in every scenario
(`r se_range("coverage", method = "delta", parameter = "auc", elimination = "slow", digits = 1)`%),
though still a few points short.  Adding the variance of $\lambda_z$ from the
individual samples
(`r se_range("coverage", method = "individual", parameter = "auc", elimination = "slow", digits = 1)`%)
recovered only part of the shortfall, because it omits the covariance of
$\hat{\lambda}_z$ with the means.  The results were alike for the oral and IV
bolus routes, so back-extrapolating $C_0$ added little uncertainty, and alike
for the one- and two-compartment models and the two designs.

**Accuracy of the AUC to infinity.**  The bias of the AUC to infinity was
`r se_range("relative_bias", method = "yuan", parameter = "auc", digits = 1)`%:
slightly high with fast elimination, where the linear trapezoidal rule
overestimates the area under the declining curve, and near zero with slow
elimination (slightly low with two compartments and BLQ samples).  The median absolute errors
were similar for the two estimates of $\lambda_z$, but the log-mean
$\lambda_z$ occasionally came out near zero with slow elimination and BLQ
samples, which made the AUC to infinity heavy-tailed (RMSE up to
`r round(max(sim_rows(results, list(method = "yuan", parameter = "auc"))$rmse))`%).
The Tobit fit uses every terminal sample, including the censored ones, rather
than means that count BLQ samples as zero, and it was steadier (RMSE up to
`r round(max(sim_rows(results, list(method = "individual", parameter = "auc"))$rmse))`%).

**The AUMC to infinity.**  All three methods under-covered the AUMC in every
scenario (delta `r se_range("coverage", method = "delta", parameter = "aumc")`%,
Yuan `r se_range("coverage", method = "yuan", parameter = "aumc")`%, individual
`r se_range("coverage", method = "individual", parameter = "aumc")`%), most
with slow elimination.  Its extrapolation is proportional to $1/\lambda_z^2$,
so the first-order delta method is a poorer approximation, and the estimate is
more heavy-tailed:  its median absolute error was
`r se_range("median_absolute_error", method = "yuan", parameter = "aumc")`%,
and with two compartments and slow elimination it was biased low
(`r se_range("relative_bias", method = "yuan", parameter = "aumc", compartments = 2, elimination = "slow")`%),
because the arithmetic mean of animals with different elimination rates
declines ever more slowly, which a single exponential from $t_{last}$ misses.

**Interval width.**  The delta-method intervals were wider than Yuan's, most
when elimination was slow; the extra width is what improved the coverage.

**The bootstrap.**  With three or four animals at each time, the bootstrap
intervals were too narrow.  For the AUC to infinity, the percentile interval
covered
`r boot_range("coverage_percentile", parameter = "auc", elimination = "fast", digits = 1)`%
with fast elimination and
`r boot_range("coverage_percentile", parameter = "auc", elimination = "slow", digits = 1)`%
with slow elimination, against
`r se_range("coverage", method = "delta", parameter = "auc", elimination = "fast", digits = 1)`%
and
`r se_range("coverage", method = "delta", parameter = "auc", elimination = "slow", digits = 1)`%
for the delta method, and the interval from the bootstrap standard error was
little better
(`r boot_range("coverage_normal", parameter = "auc", digits = 1)`%).
Resampling $n$ animals understates the variance of their mean by the factor
$(n - 1)/n$ (3/4 with four animals), and the percentile interval of a skewed
estimate is narrow as well.  For the AUMC to infinity, the bootstrap covered
`r boot_range("coverage_percentile", parameter = "aumc", digits = 1)`%:  better
than Yuan's method when elimination was slow, and as good as or better than
the delta method only with slow elimination and BLQ samples, where the
first-order delta method is weakest.  The bootstrap's advantage is that it
applies to every parameter and to ratios between groups (Shen and Machado
2017), including parameters with no analytical standard error, such as Cmax and
the half-life; with few animals at each time, its intervals should be read as
too narrow.

**More animals at each time.**  The bootstrap improved with more animals, as
its $(n - 1)/n$ shortfall shrank.  Without BLQ samples, its percentile interval
for the AUC to infinity covered
`r profile_value("coverage", 4, "none", "percentile", "auc")`% with 4 animals
per time and `r profile_value("coverage", 20, "none", "percentile", "auc")`%
with 20, when the delta method covered
`r profile_value("coverage", 20, "none", "delta", "auc")`%; with BLQ samples, it
matched or passed the delta method from 6 animals per time.  Its standard
error was then as accurate as the delta method's.  But with many animals,
every interval under-covered:  with 100 animals per time, the percentile
interval covered
`r profile_value("coverage", 100, "none", "percentile", "auc")`% without and
`r profile_value("coverage", 100, "moderate", "percentile", "auc")`% with BLQ
samples (the delta method,
`r profile_value("coverage", 100, "none", "delta", "auc")`% and
`r profile_value("coverage", 100, "moderate", "delta", "auc")`%).  The cause is
the estimate, not its standard error:  the sparse AUC to infinity was biased by
`r profile_value("median_bias", 100, "none", "yuan", "auc")`% without and
`r profile_value("median_bias", 100, "moderate", "yuan", "auc")`% with BLQ
samples, whatever the number of animals, from the BLQ samples counted as zero
in the late means and the extrapolation of a mean of animals with different
elimination rates.  More animals make every interval narrower around the same
biased estimate, so no choice of interval corrects it.  The AUMC to infinity,
biased by
`r profile_value("median_bias", 100, "none", "yuan", "aumc")`% and
`r profile_value("median_bias", 100, "moderate", "yuan", "aumc")`%, showed the
same pattern sooner.

# Arithmetic or geometric mean profiles

The sparse AUC uses the arithmetic mean at each time, whose profile is the
population mean concentration curve.  A geometric mean profile instead
approximates the typical animal, and it may be less affected by animals with
high concentrations.  The second study compared the two for the AUClast,
AUCinf, Cmax ($C_0$ for IV bolus), AUMClast, AUMCinf, and half-life, using the
same design with `r format(mean_type$n_replicates, big.mark = ",")` replicates
per scenario (`data-raw/sparse_mean_type_simulation.R`).

For each replicate, both mean profiles were formed, and `pk.nca()` calculated
the parameters from each with the linear trapezoidal rule (imputing zero at
time 0 for oral dosing and back-extrapolating $C_0$ for IV bolus).  For the
geometric mean, a time with more than half of its samples BLQ was BLQ (as for
the arithmetic sparse mean), and otherwise BLQ samples counted as half of the
LLOQ.  Each profile was compared with its own target (the arithmetic or
geometric mean of the individual parameters of
`r format(mean_type$n_virtual, big.mark = ",")` virtual animals) and with the
typical animal.  Precision is the robust coefficient of variation (the
interquartile range divided by 1.349, relative to the median).

```{r mean-type-summary, echo = FALSE}
mt <- mean_type$summary
parameter_labels <-
  c(auclast = "AUClast", aucinf = "AUCinf", cmax = "Cmax (C0 for IV)",
    aumclast = "AUMClast", aumcinf = "AUMCinf", half.life = "Half-life")
range_text <- function(x) paste(round(range(x)), collapse = " to ")
out <- data.frame()
for (current_parameter in names(parameter_labels)) {
  a <- mt[mt$parameter == current_parameter & mt$mean_type == "arithmetic", ]
  g <- mt[mt$parameter == current_parameter & mt$mean_type == "geometric", ]
  g <- g[match(a$scenario, g$scenario), ]
  out <-
    rbind(
      out,
      data.frame(
        Parameter = parameter_labels[[current_parameter]],
        `Bias vs own target, arith.` = range_text(a$bias_natural),
        `Bias vs own target, geom.` = range_text(g$bias_natural),
        `Bias vs typical, arith.` = range_text(a$bias_typical),
        `Bias vs typical, geom.` = range_text(g$bias_typical),
        `Robust CV, arith.` = range_text(a$robust_cv),
        `Robust CV, geom.` = range_text(g$robust_cv),
        `Geom. CV lower (scenarios)` = sprintf("%d of %d", sum(g$robust_cv < a$robust_cv - 1), nrow(a)),
        check.names = FALSE
      )
    )
}
knitr::kable(out, row.names = FALSE, caption = "Bias (median, %) and robust CV (%) over the scenarios; the last column counts the scenarios where the geometric robust CV was more than 1 point lower.")
```

```{r mean-type-half-life, echo = FALSE}
hl <- mt[mt$parameter == "half.life", ]
keys <- unique(hl[, c("compartments", "elimination", "blq")])
keys <- keys[order(keys$compartments, keys$elimination, keys$blq), ]
out_hl <- data.frame()
for (i in seq_len(nrow(keys))) {
  current <- merge(hl, keys[i, ])
  out_hl <-
    rbind(
      out_hl,
      data.frame(
        Model = paste0(keys$compartments[i], "-cmt"),
        Elimination = keys$elimination[i],
        BLQ = keys$blq[i],
        `Arithmetic` = range_text(current$bias_typical[current$mean_type == "arithmetic"]),
        `Geometric` = range_text(current$bias_typical[current$mean_type == "geometric"]),
        check.names = FALSE
      )
    )
}
knitr::kable(out_hl, row.names = FALSE, caption = "Bias (%) of the half-life of the mean profile against the typical animal's half-life, over routes and designs.")
```

**Findings.**  Against their own targets, the two profiles were similarly
accurate for the AUCs and Cmax.  The geometric profile was as precise or more
precise, most clearly for the half-life with BLQ samples and for the AUMC to
infinity, but the gains were modest elsewhere.  Both underestimated the AUMC to
infinity (arithmetic
`r mt_range("bias_natural", mean_type = "arithmetic", parameter = "aumcinf")`%,
geometric `r mt_range("bias_natural", mean_type = "geometric", parameter = "aumcinf")`%).
The half-life of the arithmetic mean profile was not consistently too short:
it was too long when the terminal phase was well sampled, because the late mean
is dominated by the slowly eliminating animals, and too short with BLQ samples
(counted as zero in the late means) and in the two-compartment model with slow
elimination, where the fitted points reach back toward the distribution phase.
The geometric profile was generally a few percent short, because the
geometric mean of exponentials declines at the average elimination rate.

The geometric profile was not adopted.  Its gains were modest; it estimates a
different quantity (the typical animal rather than the population mean curve);
its BLQ rule (here, half the LLOQ) is a choice that drives its results with BLQ
samples; and the sparse variance methods (Bailer, Nedelman and Jia, Holder)
are for arithmetic means, so adopting it would change every sparse calculation.

# Calculating it with PKNCA

Request `aucinf.obs` (or `aumcinf.obs`) with sparse data.  The standard error
and degrees of freedom are reported with it as `aucinf.obs_se` and
`aucinf.obs_df`, and the summary shows the estimate with its standard error.
These animals were not sampled at the dose, so the concentration at time 0 is
imputed as zero.

```{r pknca-example}
d_sparse <-
  data.frame(
    time = rep(c(0.5, 1, 2, 4, 6, 8, 12, 24), each = 4),
    conc =
      c(3.1, 4.4, 3.6, 5.2,  5.8, 6.9, 5.1, 7.4,  7.2, 6.1, 8.3, 6.6,
        5.0, 6.2, 4.4, 5.6,  4.1, 3.4, 4.8, 3.7,  2.9, 3.5, 2.4, 3.1,
        1.6, 1.9, 1.3, 1.8,  0.33, 0.27, 0.41, 0.30)
  )
d_sparse$animal <- seq_len(nrow(d_sparse))
o_data <-
  PKNCAdata(
    PKNCAconc(d_sparse, conc ~ time | animal, sparse = TRUE),
    intervals = data.frame(start = 0, end = Inf, aucinf.obs = TRUE, aumcinf.obs = TRUE),
    impute = "start_conc0"
  )
o_nca <- pk.nca(o_data)
d_result <- as.data.frame(o_nca)
d_result[grepl("^au.*inf|^lambda.z$", d_result$PPTESTCD), c("PPTESTCD", "PPORRES")]
summary(o_nca)
```

The `sparse_lambda_z_se` option (set globally with `PKNCA.options()` or for
one analysis with the `options` argument of `PKNCAdata()`) is `"delta"` (the
default) for the delta method or `"none"` for Yuan's method.  The method used
is recorded in the `PPANMETH` column.

After an IV bolus, request the IV parameters; $C_0$ is back-extrapolated from
the mean profile, and no imputation is needed.

```{r pknca-example-iv}
d_sparse_iv <-
  data.frame(
    time = rep(c(0.25, 0.5, 1, 2, 4, 8, 24), each = 4),
    conc =
      c(9.1, 10.4, 8.7, 9.9,  8.8, 9.5, 8.1, 9.2,  8.2, 7.6, 8.9, 7.9,
        6.4, 7.1, 6.0, 6.9,  4.6, 4.1, 5.0, 4.3,  2.1, 1.8, 2.3, 1.9,
        0.10, 0.08, 0.12, 0.09)
  )
d_sparse_iv$animal <- seq_len(nrow(d_sparse_iv))
o_data_iv <-
  PKNCAdata(
    PKNCAconc(d_sparse_iv, conc ~ time | animal, sparse = TRUE),
    PKNCAdose(data.frame(time = 0, dose = 100), dose ~ time, route = "intravascular"),
    intervals = data.frame(start = 0, end = Inf, c0 = TRUE, aucivinf.obs = TRUE)
  )
d_result_iv <- as.data.frame(suppressWarnings(pk.nca(o_data_iv)))
d_result_iv[grepl("^c0$|^aucivinf", d_result_iv$PPTESTCD), c("PPTESTCD", "PPORRES")]
```

The warnings suppressed here say that the AUCs from time 0 without $C_0$
(`aucinf.obs`, which `aucivinf.obs` depends on) are not calculated.

# Discussion

When $\lambda_z$ is estimated from the same samples, its uncertainty adds to
that of the AUC to infinity, and treating it as known, as Yuan's (1993)
standard error does, under-covers once the extrapolated area is a substantial
part of the AUC.  The delta method on the log means is a consistent extension of
Yuan's method:  it keeps his estimate, applies to serial-sacrifice and batch
designs alike through the covariance of the means, includes the covariance of
$\hat{\lambda}_z$ with $C_{last}$ and the AUC to $t_{last}$, and needs nothing
beyond the quantities already used for the sparse AUC to $t_{last}$.  In the
simulation it brought the coverage of the AUC to infinity much closer to
nominal when the extrapolated area was large
(`r se_range("coverage", method = "delta", parameter = "auc", elimination = "slow", digits = 1)`%,
against `r se_range("coverage", method = "yuan", parameter = "auc", elimination = "slow", digits = 1)`%
for Yuan), though still a few points short, and it cost little when the
extrapolated area was small.  For this reason, it is PKNCA's default
(`sparse_lambda_z_se = "delta"`).  The same delta method carries the
uncertainty of the back-extrapolated $C_0$ after an IV bolus, and the IV bolus
scenarios covered as well as the oral ones.

Estimating $\lambda_z$ from the individual samples with a (mixed-effects) Tobit
regression gave a more stable point estimate when samples were BLQ, but adding
its variance to Yuan's did not give adequate coverage, and its mixed-effects
fit was itself occasionally unstable in the batch design with BLQ samples and
slow elimination.  A natural refinement, not studied
here, would combine the individual-sample $\lambda_z$ with a variance that
includes its covariance with the means, for example by a joint delta method or a
bootstrap over animals.

Limitations remain.  The delta method conditions on the selected half-life
points, so the uncertainty of the automatic point selection is not included; it
is a first-order approximation, which is least accurate when $\lambda_z$ is
imprecise, as for the AUMC with slow elimination; and the arithmetic mean
profile of animals with different elimination rates is not monoexponential, so
the extrapolation $C_{last}/\lambda_z$ is itself biased for the population mean
curve, which is part of why the AUMC to infinity is underestimated.  Where the
extrapolated area is large, sampling longer remains the better remedy.

# References

Bailer AJ.  Testing for the equality of area under the curves when using
destructive measurement techniques.  Journal of Pharmacokinetics and
Biopharmaceutics.  1988;16(3):303-309.

Holder DJ.  Comments on Nedelman and Jia's extension of Satterthwaite's
approximation applied to pharmacokinetics.  Journal of Biopharmaceutical
Statistics.  2001;11(1-2):75-79.  doi:10.1081/BIP-100104199

Nedelman JR, Gibiansky E, Lau DTW.  Applying Bailer's method for AUC
confidence intervals to sparse sampling.  Pharmaceutical Research.
1995;12(1):124-128.

Nedelman JR, Jia X.  An extension of Satterthwaite's approximation applied to
pharmacokinetics.  Journal of Biopharmaceutical Statistics.
1998;8(2):317-328.  doi:10.1080/10543409808835241

Shen M, Machado SG.  Bioequivalence evaluation of sparse sampling
pharmacokinetics data using bootstrap resampling method.  Journal of
Biopharmaceutical Statistics.  2017;27(2):257-264.
doi:10.1080/10543406.2016.1265543

Takemoto S, Yamaoka K, Nishikawa M, Takakura Y.  Histogram analysis of
pharmacokinetic parameters by bootstrap resampling from one-point sampling data
in animal experiments.  Drug Metabolism and Pharmacokinetics.
2006;21(6):458-464.  doi:10.2133/dmpk.21.458

Yuan J.  Estimation of variance for AUC in animal studies.  Journal of
Pharmaceutical Sciences.  1993;82(7):761-763.  doi:10.1002/jps.2600820718
