---
title: "Bioequivalence Calculations"
author: "Bill Denney"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Bioequivalence Calculations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(PKNCA)

# The bioequivalence functions rely on packages that are Suggested (optional)
# rather than required.  The analysis chunks below only run when they are
# available.
has_be_pkgs <-
  requireNamespace("lme4", quietly = TRUE) &&
  requireNamespace("lmerTest", quietly = TRUE) &&
  requireNamespace("emmeans", quietly = TRUE)
```

After noncompartmental analysis (NCA) parameters such as the area under the
curve (AUC) and maximum concentration (C~max~) have been calculated, average
bioequivalence (BE) of a test formulation relative to a reference formulation is
assessed with inferential statistics.  PKNCA provides three functions for this
step:

* `fitbe_models()` fits a linear mixed-effects model to the log-transformed value
  of each NCA endpoint,
* `fitbe_table()` summarizes those fits into the geometric mean ratio (GMR), its
  confidence interval, and the intra-subject coefficient of variation, and
* `fitbe_calculate()` runs both in a single step.

These functions consume the results of `pk.nca()` directly, so the same workflow
that produces NCA parameters flows into the bioequivalence assessment.  For
reference-scaled assessments and regulatory pass/fail decisions (EMA, Health
Canada, GCC, and the FDA RSABE/NTID frameworks), see
[Reference scaling and regulatory frameworks](#reference-scaling-and-regulatory-frameworks)
below and `be_assess()`.

## Optional packages

The bioequivalence calculations use [`lme4`](https://cran.r-project.org/package=lme4),
[`lmerTest`](https://cran.r-project.org/package=lmerTest), and
[`emmeans`](https://cran.r-project.org/package=emmeans) to fit the mixed model
and estimate least-squares means.  They are *Suggested* packages, so install them
if they are not already present:

```{r install, eval = FALSE}
install.packages(c("lme4", "lmerTest", "emmeans"))
```

```{r no-pkgs, echo = FALSE, results = "asis", eval = !has_be_pkgs}
cat("> **Note:** `lme4`, `lmerTest`, and/or `emmeans` are not installed, so the",
    "analysis output below is not shown.  Install them to run this vignette.\n")
```

## Simulating a 2x2 crossover study

A classic average-bioequivalence study uses a two-sequence, two-period crossover
design: each subject receives both the test (T) and reference (R) formulations,
in an order determined by their randomized sequence (RT or TR).  We simulate
concentration-time data from a one-compartment model with first-order
absorption, with between-subject variability on clearance and a small difference
between formulations.

```{r simulate}
simulate_crossover <- function(nsub = 24, seed = 20251201) {
  set.seed(seed)
  times <- c(0, 0.25, 0.5, 1, 1.5, 2, 3, 4, 6, 8, 12, 16, 24)
  sequence <- rep(c("RT", "TR"), length.out = nsub)
  subj_cl <- stats::rnorm(nsub, mean = log(5), sd = 0.25)
  dose <- 100
  v <- 50
  ka <- 1.1
  rows <- list()
  for (i in seq_len(nsub)) {
    forms <- if (sequence[i] == "RT") c("R", "T") else c("T", "R")
    for (p in 1:2) {
      form <- forms[p]
      # The test formulation has slightly higher exposure (lower clearance)
      cl <- exp(subj_cl[i] + ifelse(form == "T", -0.05, 0) + stats::rnorm(1, sd = 0.06))
      ke <- cl / v
      conc <- (dose * ka / (v * (ka - ke))) * (exp(-ke * times) - exp(-ka * times))
      conc <- conc * exp(stats::rnorm(length(times), sd = 0.05))
      conc[times == 0] <- 0
      rows[[length(rows) + 1]] <-
        data.frame(
          subject = i, sequence = sequence[i], period = p, formulation = form,
          time = times, conc = conc
        )
    }
  }
  do.call(rbind, rows)
}

conc_data <- simulate_crossover()
head(conc_data)
```

## Noncompartmental analysis

The crossover structure is described in the grouping formula.  The grouping
variables are the sequence, period, and formulation, with the subject as the
final grouping variable.

```{r nca, message = FALSE, warning = FALSE}
dose_data <- conc_data[conc_data$time == 0, c("subject", "sequence", "period", "formulation")]
dose_data$dose <- 100
dose_data$time <- 0

o_conc <- PKNCAconc(conc_data, conc ~ time | sequence + period + formulation + subject)
o_dose <- PKNCAdose(dose_data, dose ~ time | sequence + period + formulation + subject)

intervals <-
  data.frame(start = 0, end = Inf, cmax = TRUE, auclast = TRUE, aucinf.obs = TRUE)
o_data <- PKNCAdata(o_conc, o_dose, intervals = intervals)

o_results <- pk.nca(o_data)
```

The NCA parameters are available with `summary()` and `as.data.frame()` as usual:

```{r nca-results}
knitr::kable(head(as.data.frame(o_results)), digits = 2)
```

## Bioequivalence assessment

`fitbe_calculate()` takes the `PKNCAresults` object, the column that identifies
the formulation (`reference_col`), and the value of that column that is the
reference (`reference_value`).  When the fixed and random effects are not
specified, they are derived from the grouping structure: the subject becomes the
random intercept and the remaining grouping variables (sequence, period, and
formulation) become the fixed effects.

```{r be, eval = has_be_pkgs}
be <- fitbe_calculate(
  o_results,
  endpoints = c("cmax", "auclast", "aucinf.obs"),
  reference_col = "formulation",
  reference_value = "R"
)

knitr::kable(be$table, digits = 2)
```

```{r be-method, eval = has_be_pkgs}
attr(be$table, "method")
```

## Interpreting the results

For each endpoint the table reports:

* `gm_test`, `gm_reference`: the geometric mean of the test and reference
  formulations (least-squares means back-transformed from the log scale),
* `gmr_percent`: the geometric mean ratio (test / reference), expressed as a
  percentage,
* `ci_lower`, `ci_upper`: the bounds of the confidence interval for the GMR
  (90% by default, controlled by `alpha`),
* `df`: the Satterthwaite degrees of freedom, and
* `cv_intra_percent`: the intra-subject coefficient of variation, derived from
  the residual standard deviation as `sqrt(exp(s^2) - 1) * 100`.

Two formulations are typically declared bioequivalent when the 90% confidence
interval for the geometric mean ratio falls entirely within 80.00%-125.00% for
the primary endpoints (commonly AUC and C~max~).

The model fit for each endpoint is

$$\log(\text{value}) = \text{sequence} + \text{period} + \text{formulation} + (1 \mid \text{subject})$$

fit with `lmerTest::lmer()` so that Satterthwaite degrees of freedom are
available, and the confidence interval is obtained by exponentiating the
difference in least-squares means between the test and reference formulations.

## Reference scaling and regulatory frameworks

Highly variable drugs (within-subject coefficient of variation above 30%) are
difficult to show bioequivalent against the fixed 80.00-125.00% limits, so
regulators allow the limits to be *scaled* by the within-reference variability,
estimated from a replicate design in which the reference is given more than
once.  PKNCA implements the major frameworks through `be_assess()`:

* **ABE** -- average bioequivalence, the unscaled 80.00-125.00% criterion.
* **EMA**, **Health Canada (HC)**, and **GCC** -- average bioequivalence with
  expanding limits (ABEL); the limits widen with the within-reference standard
  deviation, with framework-specific caps.
* **FDA RSABE** -- reference-scaled average bioequivalence using the linearized
  Howe/Hyslop criterion.
* **FDA NTID** and **HVNTID** -- the (highly variable) narrow therapeutic index
  drug frameworks, which add a constraint on the ratio of within-subject
  standard deviations.

All regulatory constants are built into PKNCA (see `be_regulator()`); no
external package is required.

### A replicate-design study

Reference scaling needs the reference replicated, so we simulate a four-period,
two-sequence full replicate design (`TRTR`/`RTRT`) with high within-subject
variability.

```{r simulate-replicate}
simulate_replicate <- function(nsub = 24, seed = 20251015, cv_wr = 0.45, cv_wt = 0.40) {
  set.seed(seed)
  seqs <- rep(c("TRTR", "RTRT"), length.out = nsub)
  swR <- sqrt(log(cv_wr^2 + 1))
  swT <- sqrt(log(cv_wt^2 + 1))
  subj <- stats::rnorm(nsub, sd = 0.4)
  rows <- list()
  for (i in seq_len(nsub)) {
    trt <- strsplit(seqs[i], "")[[1]]
    for (p in seq_along(trt)) {
      sw <- if (trt[p] == "R") swR else swT
      eff <- if (trt[p] == "T") log(1.05) else 0
      rows[[length(rows) + 1]] <- data.frame(
        subject = i, sequence = seqs[i], period = p, treatment = trt[p],
        PPTESTCD = "auclast",
        PPORRES = exp(log(100) + eff + subj[i] + stats::rnorm(1, sd = sw))
      )
    }
  }
  do.call(rbind, rows)
}

rep_data <- simulate_replicate()
```

`be_assess()` takes the long data (or a `PKNCAresults` object), the column that
identifies the formulation, the reference value, and the framework.  It returns
one row per endpoint and test formulation with the geometric mean ratio, the
confidence interval, the within-subject variability, the (possibly scaled)
acceptance limits or criterion, and the pass/fail decision.

```{r be-assess, eval = has_be_pkgs}
ema <- be_assess(rep_data, reference_col = "treatment", reference_value = "R",
                 endpoints = "auclast", regulator = "EMA")
knitr::kable(as.data.frame(ema), digits = 3)
```

For this dataset the within-reference CV is about 50%, so the EMA limits expand
from 80.00-125.00% toward their cap.  The FDA reference-scaled criterion instead
reports the linearized bound (`criterion`), which passes when it is at most
zero:

```{r be-assess-fda, eval = has_be_pkgs}
fda <- be_assess(rep_data, reference_col = "treatment", reference_value = "R",
                 endpoints = "auclast", regulator = "FDA")
knitr::kable(as.data.frame(fda)[, c("gmr_percent", "ci_lower", "ci_upper",
                                    "cvwr_percent", "swr", "criterion", "pass")],
             digits = 4)
```

### Comparing frameworks side by side

`be_compare()` runs the same study under several frameworks at once; frameworks
the design cannot support are skipped with a warning.

```{r be-compare, eval = has_be_pkgs}
cmp <- be_compare(rep_data, reference_col = "treatment", reference_value = "R",
                  endpoints = "auclast",
                  regulators = c("EMA", "HC", "GCC", "FDA", "NTID", "HVNTID"))
summary(cmp)
```

The pass/fail decision can differ between frameworks for the same data: the
narrow therapeutic index criterion is the strictest because it additionally
requires the conventional 90% confidence interval to lie within 80.00-125.00%.

## Reporting checklist

The summary supports common bioequivalence reporting requirements:

1. **Software and method**: recorded in the `method` attribute of the table
   (mixed-effects ANOVA via `lmerTest`/`lme4`, confidence interval by
   exponentiated differences in least-squares means via `emmeans`).
2. **Geometric means, %GMR, confidence interval, degrees of freedom, and
   intra-subject CV**: one row per endpoint in `be$table`.
3. **Individual data**: available from `as.data.frame(o_results)` for listings
   and figures.
