---
title: "Bioequivalence Calculations"
author: "Bill Denney"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Bioequivalence Calculations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(PKNCA)

# The bioequivalence functions rely on packages that are Suggested (optional)
# rather than required.  The analysis chunks below only run when they are
# available.
has_be_pkgs <-
  requireNamespace("lme4", quietly = TRUE) &&
  requireNamespace("lmerTest", quietly = TRUE) &&
  requireNamespace("emmeans", quietly = TRUE)
```

After noncompartmental analysis (NCA) parameters such as the area under the
curve (AUC) and maximum concentration (C~max~) have been calculated, average
bioequivalence (BE) of a test formulation relative to a reference formulation is
assessed with inferential statistics.  PKNCA provides three functions for this
step:

* `fitbe_models()` fits a linear mixed-effects model to the log-transformed value
  of each NCA endpoint,
* `fitbe_table()` summarizes those fits into the geometric mean ratio (GMR), its
  confidence interval, and the intra-subject coefficient of variation, and
* `fitbe_calculate()` runs both in a single step.

These functions consume the results of `pk.nca()` directly, so the same workflow
that produces NCA parameters flows into the bioequivalence assessment.

## Optional packages

The bioequivalence calculations use [`lme4`](https://cran.r-project.org/package=lme4),
[`lmerTest`](https://cran.r-project.org/package=lmerTest), and
[`emmeans`](https://cran.r-project.org/package=emmeans) to fit the mixed model
and estimate least-squares means.  They are *Suggested* packages, so install them
if they are not already present:

```{r install, eval = FALSE}
install.packages(c("lme4", "lmerTest", "emmeans"))
```

```{r no-pkgs, echo = FALSE, results = "asis", eval = !has_be_pkgs}
cat("> **Note:** `lme4`, `lmerTest`, and/or `emmeans` are not installed, so the",
    "analysis output below is not shown.  Install them to run this vignette.\n")
```

## Simulating a 2x2 crossover study

A classic average-bioequivalence study uses a two-sequence, two-period crossover
design: each subject receives both the test (T) and reference (R) formulations,
in an order determined by their randomized sequence (RT or TR).  We simulate
concentration-time data from a one-compartment model with first-order
absorption, with between-subject variability on clearance and a small difference
between formulations.

```{r simulate}
simulate_crossover <- function(nsub = 24, seed = 20251201) {
  set.seed(seed)
  times <- c(0, 0.25, 0.5, 1, 1.5, 2, 3, 4, 6, 8, 12, 16, 24)
  sequence <- rep(c("RT", "TR"), length.out = nsub)
  subj_cl <- stats::rnorm(nsub, mean = log(5), sd = 0.25)
  dose <- 100
  v <- 50
  ka <- 1.1
  rows <- list()
  for (i in seq_len(nsub)) {
    forms <- if (sequence[i] == "RT") c("R", "T") else c("T", "R")
    for (p in 1:2) {
      form <- forms[p]
      # The test formulation has slightly higher exposure (lower clearance)
      cl <- exp(subj_cl[i] + ifelse(form == "T", -0.05, 0) + stats::rnorm(1, sd = 0.06))
      ke <- cl / v
      conc <- (dose * ka / (v * (ka - ke))) * (exp(-ke * times) - exp(-ka * times))
      conc <- conc * exp(stats::rnorm(length(times), sd = 0.05))
      conc[times == 0] <- 0
      rows[[length(rows) + 1]] <-
        data.frame(
          subject = i, sequence = sequence[i], period = p, formulation = form,
          time = times, conc = conc
        )
    }
  }
  do.call(rbind, rows)
}

conc_data <- simulate_crossover()
head(conc_data)
```

## Noncompartmental analysis

The crossover structure is described in the grouping formula.  The grouping
variables are the sequence, period, and formulation, with the subject as the
final grouping variable.

```{r nca, message = FALSE, warning = FALSE}
dose_data <- conc_data[conc_data$time == 0, c("subject", "sequence", "period", "formulation")]
dose_data$dose <- 100
dose_data$time <- 0

o_conc <- PKNCAconc(conc_data, conc ~ time | sequence + period + formulation + subject)
o_dose <- PKNCAdose(dose_data, dose ~ time | sequence + period + formulation + subject)

intervals <-
  data.frame(start = 0, end = Inf, cmax = TRUE, auclast = TRUE, aucinf.obs = TRUE)
o_data <- PKNCAdata(o_conc, o_dose, intervals = intervals)

o_results <- pk.nca(o_data)
```

The NCA parameters are available with `summary()` and `as.data.frame()` as usual:

```{r nca-results}
knitr::kable(head(as.data.frame(o_results)), digits = 2)
```

## Bioequivalence assessment

`fitbe_calculate()` takes the `PKNCAresults` object, the column that identifies
the formulation (`reference_col`), and the value of that column that is the
reference (`reference_value`).  When the fixed and random effects are not
specified, they are derived from the grouping structure: the subject becomes the
random intercept and the remaining grouping variables (sequence, period, and
formulation) become the fixed effects.

```{r be, eval = has_be_pkgs}
be <- fitbe_calculate(
  o_results,
  endpoints = c("cmax", "auclast", "aucinf.obs"),
  reference_col = "formulation",
  reference_value = "R"
)

knitr::kable(be$table, digits = 2)
```

```{r be-method, eval = has_be_pkgs}
attr(be$table, "method")
```

## Interpreting the results

For each endpoint the table reports:

* `gm_test`, `gm_reference`: the geometric mean of the test and reference
  formulations (least-squares means back-transformed from the log scale),
* `gmr_percent`: the geometric mean ratio (test / reference), expressed as a
  percentage,
* `ci_lower`, `ci_upper`: the bounds of the confidence interval for the GMR
  (90% by default, controlled by `alpha`),
* `df`: the Satterthwaite degrees of freedom, and
* `cv_intra_percent`: the intra-subject coefficient of variation, derived from
  the residual standard deviation as `sqrt(exp(s^2) - 1) * 100`.

Two formulations are typically declared bioequivalent when the 90% confidence
interval for the geometric mean ratio falls entirely within 80.00%-125.00% for
the primary endpoints (commonly AUC and C~max~).

The model fit for each endpoint is

$$\log(\text{value}) = \text{sequence} + \text{period} + \text{formulation} + (1 \mid \text{subject})$$

fit with `lmerTest::lmer()` so that Satterthwaite degrees of freedom are
available, and the confidence interval is obtained by exponentiating the
difference in least-squares means between the test and reference formulations.

## Reporting checklist

The summary supports common bioequivalence reporting requirements:

1. **Software and method**: recorded in the `method` attribute of the table
   (mixed-effects ANOVA via `lmerTest`/`lme4`, confidence interval by
   exponentiated differences in least-squares means via `emmeans`).
2. **Geometric means, %GMR, confidence interval, degrees of freedom, and
   intra-subject CV**: one row per endpoint in `be$table`.
3. **Individual data**: available from `as.data.frame(o_results)` for listings
   and figures.
