---
title: "Bioequivalence Calculations"
author: "Bill Denney"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Bioequivalence Calculations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(PKNCA)

# The bioequivalence functions rely on packages that are Suggested (optional)
# rather than required.  The analysis chunks below only run when they are
# available.
has_be_pkgs <-
  requireNamespace("lme4", quietly = TRUE) &&
  requireNamespace("lmerTest", quietly = TRUE) &&
  requireNamespace("emmeans", quietly = TRUE)
```

After noncompartmental analysis (NCA) parameters such as the area under the
curve (AUC) and maximum concentration (C~max~) have been calculated, average
bioequivalence (BE) of a test formulation relative to a reference formulation is
assessed with inferential statistics.  PKNCA provides a single function for the
whole assessment:

* **`be_assess()`** takes the results of `pk.nca()` (or a tidy data.frame), a
  regulatory framework, and the reference formulation, and returns a table with
  the geometric mean ratio (GMR), its confidence interval, the within-subject
  variability, the (possibly scaled) acceptance limits or criterion, and the
  pass/fail decision for each endpoint.
* **`be_compare()`** runs the same study under several frameworks side by side.

These consume the results of `pk.nca()` directly, so the same workflow that
produces NCA parameters flows into the bioequivalence assessment.

## How the calculation works

`be_assess()` (and `be_compare()`) are thin verbs over a single calculation
path, the coordinator `be_fit_models()`, which runs these stages in order:

1. **`be_dataset()`** -- validate the data, resolve the value column, drop
   excluded rows, detect the subject/sequence/period columns, and set the
   reference as the first factor level.
2. **`be_design()`** -- classify the crossover design (2x2x2, full or partial
   replicate) and report which frameworks it can support.
3. **`be_fit_model_single()`** -- fit the average-BE model (this is the only
   place models are fit), dispatching on the model type to `be_fit_model_lmer()`,
   `be_fit_model_nlme()`, or `be_fit_model_anova()`.
4. **`be_extract_param()`** -- extract the geometric means, the GMR and its
   confidence interval, and the within-formulation standard deviations.
5. **`be_table()`** -- apply the regulatory limits or criterion and decide
   pass/fail (using the constants from `be_regulator()`).

The model type is selected automatically from the **regulator and the design**;
pass `model_type` to `be_assess()` to override it.  The FDA reference-scaled
family always uses intra-subject contrasts (`"anova"`); otherwise `be_design()`
decides -- a simple 2x2 crossover is analyzed by a fixed-effects ANOVA, while a
replicate design uses the mixed model.

| Regulator | Design required | Auto `model_type` | Notes |
|---|---|---|---|
| `ABE` | any (test & reference) | `anova` (2x2) / `lmer` (replicate) | unscaled 80.00-125.00% |
| `EMA` / `HC` / `GCC` | replicated reference | `lmer` | expanding limits (ABEL) |
| `FDA` (RSABE) | replicated reference | `anova` | intra-subject contrasts + linearized bound |
| `NTID` / `HVNTID` | full replicate (test & reference) | `anova` | adds a swT/swR ratio constraint |
| *(override)* | full replicate | `nlme` | treatment-specific variances from one mixed model |

The table also reports, on the measurement scale, the reference and test
geometric means with their 90% confidence intervals (`gm_reference`/`gm_test`
and their `_lower`/`_upper` bounds) and the endpoint `units` (taken from the
`PKNCAresults` object; `NA` when not provided, and a column because they differ
between C~max~ and AUC).

## Optional packages

The bioequivalence calculations use [`lme4`](https://cran.r-project.org/package=lme4),
[`lmerTest`](https://cran.r-project.org/package=lmerTest), and
[`emmeans`](https://cran.r-project.org/package=emmeans) to fit the mixed model
and estimate least-squares means.  They are *Suggested* packages, so install them
if they are not already present:

```{r install, eval = FALSE}
install.packages(c("lme4", "lmerTest", "emmeans"))
```

```{r no-pkgs, echo = FALSE, results = "asis", eval = !has_be_pkgs}
cat("> **Note:** `lme4`, `lmerTest`, and/or `emmeans` are not installed, so the",
    "analysis output below is not shown.  Install them to run this vignette.\n")
```

## Simulating a 2x2 crossover study

A classic average-bioequivalence study uses a two-sequence, two-period crossover
design: each subject receives both the test (T) and reference (R) formulations,
in an order determined by their randomized sequence (RT or TR).  We simulate
concentration-time data from a one-compartment model with first-order
absorption, with between-subject variability on clearance and a small difference
between formulations.

```{r simulate}
simulate_crossover <- function(nsub = 24, seed = 20251201) {
  set.seed(seed)
  times <- c(0, 0.25, 0.5, 1, 1.5, 2, 3, 4, 6, 8, 12, 16, 24)
  sequence <- rep(c("RT", "TR"), length.out = nsub)
  subj_cl <- stats::rnorm(nsub, mean = log(5), sd = 0.25)
  dose <- 100
  v <- 50
  ka <- 1.1
  rows <- list()
  for (i in seq_len(nsub)) {
    forms <- if (sequence[i] == "RT") c("R", "T") else c("T", "R")
    for (p in 1:2) {
      form <- forms[p]
      # The test formulation has slightly higher exposure (lower clearance)
      cl <- exp(subj_cl[i] + ifelse(form == "T", -0.05, 0) + stats::rnorm(1, sd = 0.06))
      ke <- cl / v
      conc <- (dose * ka / (v * (ka - ke))) * (exp(-ke * times) - exp(-ka * times))
      conc <- conc * exp(stats::rnorm(length(times), sd = 0.05))
      conc[times == 0] <- 0
      rows[[length(rows) + 1]] <-
        data.frame(
          subject = i, sequence = sequence[i], period = p, formulation = form,
          time = times, conc = conc
        )
    }
  }
  do.call(rbind, rows)
}

conc_data <- simulate_crossover()
head(conc_data)
```

## Noncompartmental analysis

The crossover structure is described in the grouping formula.  The grouping
variables are the sequence, period, and formulation, with the subject as the
final grouping variable.

```{r nca, message = FALSE, warning = FALSE}
dose_data <- conc_data[conc_data$time == 0, c("subject", "sequence", "period", "formulation")]
dose_data$dose <- 100
dose_data$time <- 0

o_conc <- PKNCAconc(conc_data, conc ~ time | sequence + period + formulation + subject)
o_dose <- PKNCAdose(dose_data, dose ~ time | sequence + period + formulation + subject)

intervals <-
  data.frame(start = 0, end = Inf, cmax = TRUE, auclast = TRUE, aucinf.obs = TRUE)
o_data <- PKNCAdata(o_conc, o_dose, intervals = intervals)

o_results <- pk.nca(o_data)
```

The NCA parameters are available with `summary()` and `as.data.frame()` as usual:

```{r nca-results}
knitr::kable(head(as.data.frame(o_results)), digits = 2)
```

## Bioequivalence assessment

`be_assess()` takes the `PKNCAresults` object, the column that identifies the
formulation (`reference_col`), the value of that column that is the reference
(`reference_value`), and the regulatory framework.  This 2x2 study is not
replicated, so only average bioequivalence (`regulator = "ABE"`) is feasible.

```{r be, eval = has_be_pkgs}
be <- be_assess(
  o_results,
  endpoints = c("cmax", "auclast", "aucinf.obs"),
  reference_col = "formulation",
  reference_value = "R",
  regulator = "ABE"
)

knitr::kable(as.data.frame(be), digits = 2)
```

## Interpreting the results

For each endpoint and test formulation the table reports:

* `units`: the measurement units of the endpoint (e.g. `ng/mL` for C~max~,
  `hr*ng/mL` for AUC), taken from the `PKNCAresults` object,
* `gm_reference`, `gm_test` (with `_lower`/`_upper` bounds): the reference and
  test geometric means and their 90% confidence intervals, on the measurement
  scale (in `units`),
* `gmr_percent`: the geometric mean ratio (test / reference), as a percentage,
* `ci_lower`, `ci_upper`: the bounds of the confidence interval for the GMR
  (90% by default, controlled by `alpha`),
* `cvwr_percent`, `swr`: the within-reference coefficient of variation and
  standard deviation (used by the reference-scaling frameworks; `NA` for
  unscaled ABE),
* `limit_lower`, `limit_upper`: the acceptance limits, and `criterion` for the
  FDA reference-scaled bound, and
* `pass`: the regulatory decision.

Two formulations are declared bioequivalent under ABE when the 90% confidence
interval for the geometric mean ratio falls entirely within 80.00%-125.00% for
the primary endpoints (commonly AUC and C~max~).

For ABE the average-BE model fit for each endpoint is

$$\log(\text{value}) = \text{sequence} + \text{period} + \text{formulation} + (1 \mid \text{subject})$$

fit with `lmerTest::lmer()` so that Satterthwaite degrees of freedom are
available, and the confidence interval is obtained by exponentiating the
difference in least-squares means between the test and reference formulations.

## Reference scaling and regulatory frameworks

Highly variable drugs (within-subject coefficient of variation above 30%) are
difficult to show bioequivalent against the fixed 80.00-125.00% limits, so
regulators allow the limits to be *scaled* by the within-reference variability,
estimated from a replicate design in which the reference is given more than
once.  PKNCA implements the major frameworks through `be_assess()`:

* **ABE** -- average bioequivalence, the unscaled 80.00-125.00% criterion.
* **EMA**, **Health Canada (HC)**, and **GCC** -- average bioequivalence with
  expanding limits (ABEL); the limits widen with the within-reference standard
  deviation, with framework-specific caps.
* **FDA RSABE** -- reference-scaled average bioequivalence using the linearized
  Howe/Hyslop criterion.
* **FDA NTID** and **HVNTID** -- the (highly variable) narrow therapeutic index
  drug frameworks, which add a constraint on the ratio of within-subject
  standard deviations.

All regulatory constants are built into PKNCA (see `be_regulator()`); no
external package is required.

### A replicate-design study

Reference scaling needs the reference replicated, so we simulate a four-period,
two-sequence full replicate design (`TRTR`/`RTRT`) with high within-subject
variability.

```{r simulate-replicate}
simulate_replicate <- function(nsub = 24, seed = 20251015, cv_wr = 0.45, cv_wt = 0.40) {
  set.seed(seed)
  seqs <- rep(c("TRTR", "RTRT"), length.out = nsub)
  swR <- sqrt(log(cv_wr^2 + 1))
  swT <- sqrt(log(cv_wt^2 + 1))
  subj <- stats::rnorm(nsub, sd = 0.4)
  rows <- list()
  for (i in seq_len(nsub)) {
    trt <- strsplit(seqs[i], "")[[1]]
    for (p in seq_along(trt)) {
      sw <- if (trt[p] == "R") swR else swT
      eff <- if (trt[p] == "T") log(1.05) else 0
      rows[[length(rows) + 1]] <- data.frame(
        subject = i, sequence = seqs[i], period = p, treatment = trt[p],
        PPTESTCD = "auclast",
        PPORRES = exp(log(100) + eff + subj[i] + stats::rnorm(1, sd = sw))
      )
    }
  }
  do.call(rbind, rows)
}

rep_data <- simulate_replicate()
```

`be_assess()` takes the long data (or a `PKNCAresults` object), the column that
identifies the formulation, the reference value, and the framework.  It returns
one row per endpoint and test formulation with the geometric mean ratio, the
confidence interval, the within-subject variability, the (possibly scaled)
acceptance limits or criterion, and the pass/fail decision.

```{r be-assess, eval = has_be_pkgs}
ema <- be_assess(rep_data, reference_col = "treatment", reference_value = "R",
                 endpoints = "auclast", regulator = "EMA")
knitr::kable(as.data.frame(ema), digits = 3)
```

For this dataset the within-reference CV is about 50%, so the EMA limits expand
from 80.00-125.00% toward their cap.  The FDA reference-scaled criterion instead
reports the linearized bound (`criterion`), which passes when it is at most
zero:

```{r be-assess-fda, eval = has_be_pkgs}
fda <- be_assess(rep_data, reference_col = "treatment", reference_value = "R",
                 endpoints = "auclast", regulator = "FDA")
knitr::kable(as.data.frame(fda)[, c("gmr_percent", "ci_lower", "ci_upper",
                                    "cvwr_percent", "swr", "criterion", "pass")],
             digits = 4)
```

### Comparing frameworks side by side

`be_compare()` runs the same study under several frameworks at once; frameworks
the design cannot support are skipped with a warning.

```{r be-compare, eval = has_be_pkgs}
cmp <- be_compare(rep_data, reference_col = "treatment", reference_value = "R",
                  endpoints = "auclast",
                  regulators = c("EMA", "HC", "GCC", "FDA", "NTID", "HVNTID"))
summary(cmp)
```

The pass/fail decision can differ between frameworks for the same data: the
narrow therapeutic index criterion is the strictest because it additionally
requires the conventional 90% confidence interval to lie within 80.00-125.00%.

## Reporting checklist

The assessment table supports common bioequivalence reporting requirements:

1. **Framework and model**: the `regulator` and `model_type` columns record the
   framework and the average-BE model used, and the table's `caption` attribute
   gives a one-sentence methods statement (model, confidence-interval method,
   within-subject variability, and the decision rule).
2. **%GMR, confidence interval, within-subject CV, limits/criterion, and the
   pass/fail decision**: one row per endpoint and test formulation in
   `as.data.frame(be_assess(...))`.
3. **Cross-regulator comparison**: `summary(be_compare(...))` gives an
   endpoint-by-regulator pass/fail grid.
4. **Individual data**: available from `as.data.frame(o_results)` for listings
   and figures.
